---
title: 'Greenhouse gas emissions in context: Raw totals, income groups and human development'
author: ''
date: '2026-09-01'
slug: greenhouse-gas-emissions-in-context
code_folding: hide
categories:
  - research methods
tags:
  - s
  - visualisation
  - data science
  - open science
subtitle: ''
summary: 'Greenhouse gas emissions can be reported as raw totals or per-capita values, each answering a different descriptive question. Using the 2025 EDGAR release, the World Bank income classification and the Human Development Index, this post compares totals, per-capita figures, sectors and a descriptive association with human development.'
abstract: 'This article analyses 2024 greenhouse gas emissions from the 2025 EDGAR release, separating total from per-capita emissions and reading both alongside income group and human development. It documents the data preparation, visualises regional totals, the largest emitters and their sectors, and reports the residual of each country from a descriptive association with the Human Development Index. None of these quantities estimates responsibility, causation or feasible mitigation.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
# Swap in the smallest sufficient plotly build, keeping the full one if the
# download fails, so that a render without network access still succeeds.
lighter_bundle <- function(p) {
  tryCatch(plotly::partial_bundle(p), error = function(e) {
    message('Keeping the full plotly bundle: ', conditionMessage(e))
    p
  })
}

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, fig.align = 'center',
                      comment = '#>', fig.width = 8, fig.retina = 2)
```

## Introduction

National greenhouse gas figures can be reported as totals, per-capita values, sector compositions or associations with measures of development, and each of these descriptions answers a different question. None of them, on its own, measures historical responsibility, establishes causation or determines a fair mitigation burden.

This post uses the 2025 report of the Emissions Database for Global Atmospheric Research (EDGAR), a consistent inventory of emissions by country and sector, to describe emissions in 2024 (Crippa et al., 2025). The analysis combines those estimates with the World Bank income classification and with the Human Development Index (HDI) of the United Nations Development Programme. Its aim is to make the denominators and contextual variables explicit.

## Development and plots

The figures move from totals to per-capita emissions by income group, then to sector composition and finally to a descriptive association with human development, in which the HDI serves as a contextual covariate. Each figure caption gives the source and the unit.


### Data and preparation

All emissions come from the 2025 EDGAR booklet (Crippa et al., 2025), which runs from 1970 to 2024. Greenhouse gas (GHG) here covers fossil CO<sub>2</sub>, CH<sub>4</sub>, N<sub>2</sub>O and fluorinated gases, aggregated with the 100-year global warming potentials of the IPCC Fifth Assessment Report, and excludes land use, land-use change and forestry. The booklet reports totals in million tonnes of CO<sub>2</sub> equivalent per year (Mt CO<sub>2</sub>eq) and per-capita emissions in tonnes per person. It also includes the EU27 and the world as rows of their own, and their totals are used as reported. The world population and the world per-capita figure are instead recomputed from the territories. This puts the world per-capita figure slightly below EDGAR's own, because the emissions of international aviation and shipping drop out of the numerator while the population stays the same.

The analysis reads three compact files that were derived from the booklet and from three context sources in a preparation step described at the end of this post. `ghg_long.csv` has one row per territory and year, with its continent, income group, total, per-capita figure, derived population and euro-area membership. `hdi_2023.csv` gives the Human Development Index for 2023 from the 2025 Human Development Report (United Nations Development Programme, 2025), and `sectors_2024.csv` gives the 2024 emissions of each territory by EDGAR sector, summed over the four substances. Continents come from the ISO 3166 dataset maintained by Duncalfe (n.d.) and income groups from the World Bank classification for fiscal year 2026 (World Bank, 2025).

The population of each territory and year is the total divided by the per-capita figure, which recovers the population series behind EDGAR's own per-capita values without a further source. EDGAR reports Serbia and Montenegro (code SCG) and Sudan and South Sudan (code SDN) as combined territories. It likewise folds Liechtenstein into Switzerland, Andorra into Spain, Monaco into France, San Marino and the Holy See into Italy, and the State of Palestine into Israel. These rows are kept exactly as reported and are later compared with the HDI of the member whose code they carry, except Serbia and Montenegro, whose code belongs to neither member and is matched to Serbia by hand. The first chunk loads the packages, sets one theme for all static figures, and fixes the colours, taken from the Okabe-Ito palette, that each continent and income group keeps throughout.

```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(scales)
library(ggrepel)
library(ggtext)
library(plotly)
library(htmltools)
library(leaflet)
library(sf)
library(rnaturalearth)
library(RColorBrewer)
library(knitr)

theme_set(theme_minimal(base_size = 13))
theme_update(
  # Set the leaf elements, not the axis.title parent: under ggplot2 4.0 the parent no
  # longer propagates the element_markdown class down to axis.title.x/axis.title.y, so
  # the axes print a literal "CO<sub>2</sub>" instead of a subscript
  axis.title.x = element_markdown(),
  axis.title.y = element_markdown(angle = 90),
  plot.title = element_text(size = 13, face = 'bold'),
  plot.title.position = 'plot',
  # element_textbox_simple wraps at the width of the figure; element_markdown
  # would run the caption off the right edge of a wide plot
  plot.caption = element_textbox_simple(hjust = 0, colour = 'grey30', size = 10,
                                        lineheight = 1.25, width = unit(1, 'npc'),
                                        margin = margin(t = 10)),
  plot.caption.position = 'plot',
  panel.grid.minor = element_blank(),
  legend.position = 'bottom'
)

# ggrepel places labels partly at random, so an unseeded knit can move the
# end-of-line labels of figures 1 and 2. Some draws overlap two labels in the
# Low income panel of figure 2. This seed was checked to keep every repelled
# label apart at the figure sizes used below.
repel_seed <- 2025

continent_colours <- c(Asia = '#D55E00', America = '#0072B2', Europe = '#009E73',
                       Africa = '#E69F00', Oceania = '#CC79A7',
                       `International transport` = '#56B4E9')
income_colours <- c(`High income` = '#D55E00', `Upper middle income` = '#E69F00',
                    `Lower middle income` = '#0072B2', `Low income` = '#009E73')

fmt <- function(x, digits = 0) formatC(x, format = 'f', digits = digits, big.mark = ',')
pct <- function(x, digits = 1) paste0(fmt(x, digits), '%')
ordinal <- label_ordinal()
edgar_caption <- 'Source: EDGAR 2025 (Crippa et al., 2025).'
caption_theme <- theme(
  plot.caption = element_textbox_simple(hjust = 0, colour = 'grey30', size = 10,
                                        lineheight = 1.25, width = unit(1, 'npc'),
                                        margin = margin(t = 10)))
```

The next chunk reads the three files and separates EDGAR's own EU27 and world rows from the territories. It then gives an income group to Curaçao and to Serbia and Montenegro, two territories that EDGAR reports under legacy codes absent from the World Bank classification. For fiscal year 2026, the World Bank classified Curaçao as high income and both Serbia and Montenegro as upper middle income (World Bank, 2025). The chunk also lists the jointly reported territories with their members, so that the map in figure 6 can colour each member.

```{r data}
ghg <- read.csv('data/ghg_long.csv', stringsAsFactors = FALSE)
hdi <- read.csv('data/hdi_2023.csv', stringsAsFactors = FALSE)
sectors_2024 <- read.csv('data/sectors_2024.csv', stringsAsFactors = FALSE)

world <- ghg %>% filter(iso3 == 'GLOBAL TOTAL') %>% select(year, world_mt = total_mt)
eu27 <- ghg %>% filter(iso3 == 'EU27') %>% select(year, eu27_mt = total_mt)
world_2024 <- world$world_mt[world$year == 2024]

# Countries, plus international aviation (AIR) and shipping (SEA)
countries <- ghg %>% filter(!iso3 %in% c('EU27', 'GLOBAL TOTAL'))

# EDGAR reports two territories under legacy codes that match no economy in the
# World Bank classification, so the join by code left them without an income
# group. ANT is the former Netherlands Antilles, which EDGAR labels Curaçao, and
# SCG is Serbia and Montenegro. In the World Bank's file for fiscal year 2026
# (CLASS_2025_07_02.xlsx), Curaçao and Sint Maarten are both high income. Serbia
# and Montenegro are both upper middle income, as is Kosovo, which has no EDGAR
# row of its own. Each code therefore takes the group that all its classified
# successors share, whichever of them its row covers.
legacy_income_groups <- c(ANT = 'High income', SCG = 'Upper middle income')
countries <- countries %>%
  mutate(income_group = coalesce(income_group, unname(legacy_income_groups[iso3])))

# Territories that EDGAR reports jointly under one code, with their members
combined <- tribble(
  ~iso3, ~member_iso3, ~member,
  'SCG', 'SRB', 'Serbia',       'SCG', 'MNE', 'Montenegro',
  'SDN', 'SDN', 'Sudan',        'SDN', 'SSD', 'South Sudan',
  'ISR', 'ISR', 'Israel',       'ISR', 'PSE', 'State of Palestine',
  'CHE', 'CHE', 'Switzerland',  'CHE', 'LIE', 'Liechtenstein',
  'ESP', 'ESP', 'Spain',        'ESP', 'AND', 'Andorra',
  'FRA', 'FRA', 'France',       'FRA', 'MCO', 'Monaco',
  'ITA', 'ITA', 'Italy',        'ITA', 'SMR', 'San Marino', 'ITA', 'VAT', 'Holy See'
)

# A row without a continent would otherwise vanish silently from figure 1
stopifnot(!anyNA(countries$continent))
c(territories = n_distinct(countries$iso3), years = n_distinct(ghg$year),
  countries_with_hdi = nrow(hdi), sector_rows = nrow(sectors_2024))
```

### World and regional totals

Figure 1 shows the world total by continent since 1970, with international aviation and shipping as a category of their own, because EDGAR attributes them to no country. The lower panel expresses the emissions of the EU27 and of the euro area as a percentage of the world total, since in absolute terms the world total would dwarf the European series.

```{r fig1-world-regions, fig.height = 9.5, fig.alt = paste0('Two panels. Panel a is a stacked area chart of world greenhouse gas emissions from 1970 to 2024 by continent, rising from about ', fmt(world$world_mt[world$year == 1970]), ' to about ', fmt(world_2024), ' million tonnes of CO2 equivalent, with Asia growing to more than half of the total. Panel b shows the EU27 and euro-area emissions as a percentage of the world total, both falling from about a fifth and a seventh in 1970 to a few per cent in 2024, with a dotted line marking the year in which EU27 emissions peaked in absolute terms.')}
by_continent <- countries %>%
  group_by(continent, year) %>%
  summarise(total_mt = sum(total_mt), .groups = 'drop')

continent_order <- by_continent %>%
  filter(year == 2024) %>%
  arrange(desc(total_mt)) %>%
  pull(continent)

by_continent <- by_continent %>%
  mutate(continent = factor(continent, levels = continent_order))

continent_share <- function(cont, yr) {
  100 * by_continent$total_mt[by_continent$continent == cont & by_continent$year == yr] /
    world$world_mt[world$year == yr]
}

# Both panels print their labels in a right margin of the same width, and both
# fix the x limits to the years of the data, which the nudged labels would
# otherwise stretch, so that the years of the two panels line up vertically
label_space <- 95
label_margin <- margin(5.5, label_space, 5.5, 5.5)
year_limits <- range(by_continent$year)

# One label per band, level with the middle of the band in the last year.
# position_stack(reverse = TRUE) stacks the first level at the bottom.
continent_labels <- by_continent %>%
  filter(year == max(year)) %>%
  arrange(continent) %>%
  mutate(label_y = cumsum(total_mt) - total_mt / 2,
         label = sub(' ', '\n', continent))

p1a <- ggplot(by_continent, aes(year, total_mt, fill = continent)) +
  geom_area(position = position_stack(reverse = TRUE), colour = 'white', linewidth = 0.3) +
  geom_text_repel(data = continent_labels, aes(year, label_y, label = label),
                  inherit.aes = FALSE, hjust = 0, nudge_x = 1.5, direction = 'y',
                  xlim = c(-Inf, Inf), ylim = c(-Inf, Inf), size = 3.8, lineheight = 0.9,
                  colour = 'grey20', segment.colour = 'grey50', segment.size = 0.3,
                  min.segment.length = 0, box.padding = 0.15, seed = repel_seed) +
  scale_fill_manual(values = continent_colours, guide = 'none') +
  scale_x_continuous(breaks = seq(1970, 2020, 10), limits = year_limits, oob = oob_keep,
                     expand = c(0, 0)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.04))) +
  coord_cartesian(clip = 'off') +
  labs(title = 'a. World total by continent',
       x = NULL, y = 'Mt CO<sub>2</sub>eq per year') +
  theme(plot.margin = label_margin)

# Shares of the world total: EU27 from EDGAR's own row, euro area summed over members
shares <- world %>%
  left_join(eu27, by = 'year') %>%
  left_join(countries %>%
              filter(euro_area) %>%
              group_by(year) %>%
              summarise(euro_area_mt = sum(total_mt), .groups = 'drop'),
            by = 'year') %>%
  mutate(EU27 = 100 * eu27_mt / world_mt,
         `Euro area` = 100 * euro_area_mt / world_mt)

eu27_peak_year <- shares$year[which.max(shares$eu27_mt)]
eu27_peak_mt <- max(shares$eu27_mt)
eu27_value <- function(yr) shares$eu27_mt[shares$year == yr]
share_value <- function(bloc, yr) shares[[bloc]][shares$year == yr]

shares_long <- shares %>%
  select(year, EU27, `Euro area`) %>%
  pivot_longer(-year, names_to = 'bloc', values_to = 'share')

p1b <- ggplot(shares_long, aes(year, share, colour = bloc)) +
  annotate('segment', x = eu27_peak_year, xend = eu27_peak_year, y = 0,
           yend = share_value('EU27', eu27_peak_year), linetype = 'dotted', colour = 'grey40') +
  annotate('text', x = eu27_peak_year + 1, y = 0.8,
           label = paste0('Peak of EU27 emissions in absolute terms:\n',
                          eu27_peak_year, ', ', fmt(eu27_peak_mt), ' Mt'),
           hjust = 0, vjust = 0, size = 3.6, colour = 'grey30', lineheight = 0.95) +
  geom_line(linewidth = 1) +
  geom_text_repel(data = filter(shares_long, year == max(year)),
                  aes(label = paste0(bloc, ', ', fmt(share, 1), '%')),
                  hjust = 0, nudge_x = 1.5, direction = 'y', segment.colour = NA,
                  xlim = c(-Inf, Inf), size = 3.8, show.legend = FALSE, seed = repel_seed) +
  scale_colour_manual(values = c(EU27 = '#0072B2', `Euro area` = '#D55E00')) +
  scale_x_continuous(breaks = seq(1970, 2020, 10), limits = year_limits, oob = oob_keep,
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, NA), labels = function(x) paste0(x, '%'),
                     expand = expansion(mult = c(0, 0.08))) +
  coord_cartesian(clip = 'off') +
  labs(title = 'b. EU27 and euro-area emissions as a share of the world total',
       x = NULL, y = 'Share of the world total') +
  theme(legend.position = 'none', plot.margin = label_margin)

p1a / p1b +
  plot_layout(heights = c(1.15, 1)) +
  plot_annotation(
    caption = paste0(edgar_caption, ' Units: million tonnes of CO<sub>2</sub> equivalent per year',
                     ' (panel a) and per cent of the world total (panel b).<br>',
                     'EU27: EDGAR\'s own row. Euro area: the sum of its 21 members as of 2026,',
                     ' including Bulgaria.'),
    # patchwork sets the caption across the panels but not the plot margins, so
    # the caption box is widened by the label margin to span the whole figure
    theme = theme(plot.caption = element_textbox_simple(
      hjust = 0, colour = 'grey30', size = 10, lineheight = 1.25, margin = margin(t = 10),
      width = unit(1, 'npc') + unit(label_space - 5.5, 'pt')))
  )
```

World emissions reached `r fmt(world_2024)` Mt CO<sub>2</sub>eq in 2024, `r pct(100 * (world_2024 / world$world_mt[world$year == 1970] - 1), 0)` more than in 1970. Asia accounted for `r pct(continent_share('Asia', 2024))` of the 2024 total, up from `r pct(continent_share('Asia', 1970))` in 1970, while the share of Europe fell from `r pct(continent_share('Europe', 1970))` to `r pct(continent_share('Europe', 2024))` and that of America from `r pct(continent_share('America', 1970))` to `r pct(continent_share('America', 2024))`. International aviation and shipping add `r pct(continent_share('International transport', 2024))`. The EU27 emitted `r fmt(eu27_value(2024))` Mt in 2024, `r pct(share_value('EU27', 2024))` of the world total against `r pct(share_value('EU27', 1970))` in 1970, and the euro area `r pct(share_value('Euro area', 2024))` against `r pct(share_value('Euro area', 1970))`. In absolute terms, EU27 emissions peaked in `r eu27_peak_year` at `r fmt(eu27_peak_mt)` Mt and stayed within `r pct(ceiling(100 * (1 - min(shares$eu27_mt[shares$year %in% 1980:1989]) / eu27_peak_mt)), 0)` of that peak throughout the 1980s. They have since fallen by `r pct(100 * (1 - eu27_value(2024) / eu27_peak_mt), 0)`, or by `r pct(100 * (1 - eu27_value(2024) / eu27_value(1990)), 0)` relative to 1990, the reference year of most climate targets. Panel (a) cannot show this decline because the world total grew by `r pct(100 * (world_2024 / world$world_mt[world$year == eu27_peak_year] - 1), 0)` over the same years.


### Income groups

Per-capita figures cannot be added across countries, so each income group is summarised by its population-weighted mean. The mean divides the total emissions of the group by its population, which gives the per-capita emission of the group treated as one territory. Panel (a) of figure 2 shows this quantity for the four World Bank income groups of fiscal year 2026 (World Bank, 2025). Panel (b) shows, for each group, its share of world emissions as a solid line and its share of world population as a dashed line. The distance between the two lines is the gap between what a group emits and what it would emit at the world average.

```{r fig2-income-groups, fig.height = 10, fig.alt = 'Two panels. Panel a shows population-weighted mean per-capita emissions from 1970 to 2024 for four World Bank income groups: the high-income line stays far above the others, changes little until the mid-2000s and then falls, the upper-middle-income line rises steeply after 2000 to more than half the high-income level, and the lower-middle-income and low-income lines stay below 3 tonnes per person. Panel b has one small panel per income group with its share of world emissions as a solid line and of world population as a dashed line: the high-income share of emissions falls from two thirds to one third while its population share falls from about a quarter to under a fifth, and in the two lowest groups the population share lies far above the emission share.'}
world_population <- countries %>%
  filter(!is.na(population)) %>%
  group_by(year) %>%
  summarise(world_population = sum(population), .groups = 'drop')

no_income_group <- countries %>%
  filter(year == 2024, !iso3 %in% c('AIR', 'SEA'), is.na(income_group))

by_income <- countries %>%
  filter(!is.na(income_group)) %>%
  group_by(income_group, year) %>%
  summarise(total_mt = sum(total_mt), population = sum(population),
            n_countries = n_distinct(iso3), .groups = 'drop') %>%
  left_join(world, by = 'year') %>%
  left_join(world_population, by = 'year') %>%
  mutate(per_capita_t = total_mt / population * 1e6,
         emission_share = 100 * total_mt / world_mt,
         population_share = 100 * population / world_population,
         income_group = factor(income_group, levels = names(income_colours)))

income_value <- function(group, yr, variable) {
  by_income[[variable]][by_income$income_group == group & by_income$year == yr]
}

p2a <- ggplot(by_income, aes(year, per_capita_t, colour = income_group)) +
  geom_line(linewidth = 1) +
  geom_text_repel(data = filter(by_income, year == max(year)),
                  aes(label = paste0(income_group, ', ', fmt(per_capita_t, 1), ' t')),
                  hjust = 0, nudge_x = 1, direction = 'y', segment.colour = NA,
                  xlim = c(-Inf, Inf), size = 3.8, show.legend = FALSE, seed = repel_seed) +
  scale_colour_manual(values = income_colours) +
  scale_x_continuous(breaks = seq(1970, 2020, 10), expand = expansion(mult = c(0, 0.38))) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(clip = 'off') +
  labs(title = 'a. Population-weighted mean per-capita emissions by World Bank income group',
       x = NULL, y = 't CO<sub>2</sub>eq per person') +
  theme(legend.position = 'none')

shares_by_income <- by_income %>%
  select(income_group, year, emission_share, population_share) %>%
  pivot_longer(c(emission_share, population_share), names_to = 'measure', values_to = 'share') %>%
  mutate(measure = factor(measure, levels = c('emission_share', 'population_share'),
                          labels = c('of emissions', 'of population')))

p2b <- ggplot(shares_by_income, aes(year, share, colour = income_group, linetype = measure)) +
  geom_line(linewidth = 0.9) +
  geom_text_repel(data = filter(shares_by_income, year == max(year)),
                  aes(label = paste0(fmt(share, 0), '% ', measure)),
                  hjust = 0, nudge_x = 1, direction = 'y', segment.colour = NA,
                  xlim = c(-Inf, Inf), size = 3.5, show.legend = FALSE, seed = repel_seed) +
  facet_wrap(~ income_group, nrow = 2) +
  scale_colour_manual(values = income_colours, guide = 'none') +
  scale_linetype_manual(values = c('solid', '22'), guide = 'none') +
  scale_x_continuous(breaks = seq(1970, 2010, 20), expand = expansion(mult = c(0, 0.56))) +
  scale_y_continuous(limits = c(0, NA), labels = function(x) paste0(x, '%'),
                     expand = expansion(mult = c(0, 0.08))) +
  coord_cartesian(clip = 'off') +
  labs(title = 'b. Share of world emissions (solid) and of world population (dashed)',
       x = NULL, y = 'Share of the world total') +
  theme(legend.position = 'none', strip.text = element_text(hjust = 0, face = 'bold'),
        panel.spacing.x = unit(1.5, 'lines'))

p2a / p2b +
  plot_layout(heights = c(1, 1.25)) +
  plot_annotation(
    caption = paste0(edgar_caption, ' Income groups: World Bank (2025), fiscal year 2026.',
                     ' Units: tonnes of CO<sub>2</sub> equivalent per person per year (panel a)',
                     ' and per cent of the world total (panel b).<br>',
                     'Population is the EDGAR total divided by the per-capita figure. ',
                     nrow(no_income_group), ' territories without an income group are omitted.'),
    theme = caption_theme
  )
```

In 2024, a resident of a high-income country emitted `r fmt(income_value('High income', 2024, 'per_capita_t'), 1)` t on average, `r fmt(income_value('High income', 2024, 'per_capita_t') / income_value('Low income', 2024, 'per_capita_t'), 1)` times the `r fmt(income_value('Low income', 2024, 'per_capita_t'), 1)` t of a resident of a low-income country. The high-income mean peaked at `r fmt(max(by_income$per_capita_t[by_income$income_group == 'High income']), 1)` t in `r by_income$year[by_income$income_group == 'High income'][which.max(by_income$per_capita_t[by_income$income_group == 'High income'])]` and has fallen since. The upper-middle-income mean, for a group that includes China, rose from `r fmt(income_value('Upper middle income', 2000, 'per_capita_t'), 1)` t in 2000 to `r fmt(income_value('Upper middle income', 2024, 'per_capita_t'), 1)` t in 2024. The group's share of world emissions overtook the high-income share in `r min(by_income$year[by_income$income_group == 'Upper middle income' & by_income$emission_share > by_income$emission_share[by_income$income_group == 'High income']])`. Panel (b) shows the imbalance between people and emissions: high-income countries hold `r pct(income_value('High income', 2024, 'population_share'))` of the world's population and emit `r pct(income_value('High income', 2024, 'emission_share'))` of the total. The lower-middle-income and low-income groups together hold `r pct(income_value('Lower middle income', 2024, 'population_share') + income_value('Low income', 2024, 'population_share'))` of the population and emit `r pct(income_value('Lower middle income', 2024, 'emission_share') + income_value('Low income', 2024, 'emission_share'))`. The `r nrow(no_income_group)` territories without an income group emitted `r pct(100 * sum(no_income_group$total_mt) / world_2024)` of the 2024 total between them. They include Ethiopia and Venezuela, both of which the World Bank left unclassified for fiscal year 2026, and small dependencies such as Guadeloupe and the Falkland Islands.


### The twenty largest emitters

Figure 3 ranks countries by their 2024 total, coloured by continent, with each country's share of the world total printed at the end of its bar and its per-capita emissions in a dot plot beside it.

```{r fig3-top-emitters, fig.height = 7, fig.alt = 'Horizontal bar chart of the twenty largest emitters in 2024, coloured by continent, with China far ahead, followed by the United States, India, Russia, Indonesia, Brazil, Japan and Iran. A dot plot beside the bars shows per-capita emissions, which are highest for Saudi Arabia, Australia, Canada, Russia and the United States and lowest for Pakistan, India and Indonesia.'}
top20 <- countries %>%
  filter(year == 2024, !iso3 %in% c('AIR', 'SEA')) %>%
  mutate(share = 100 * total_mt / world_2024) %>%
  arrange(desc(total_mt)) %>%
  slice_head(n = 20) %>%
  mutate(country = factor(country, levels = rev(country)))

top_value <- function(code, variable) top20[[variable]][top20$iso3 == code]
# The world total includes international aviation and shipping, whose two rows
# carry no population, so the reference line uses the country total over the
# country population, keeping the numerator and the denominator on the same set.
countries_2024 <- countries %>% filter(year == 2024, !iso3 %in% c('AIR', 'SEA'))
world_per_capita <- sum(countries_2024$total_mt) / sum(countries_2024$population) * 1e6

p3a <- ggplot(top20, aes(total_mt, country, fill = continent)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = pct(share)), hjust = -0.15, size = 3.5, colour = 'grey20') +
  scale_fill_manual(values = continent_colours, breaks = names(continent_colours)) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.16))) +
  labs(title = 'a. Total emissions and share of the world total',
       x = 'Mt CO<sub>2</sub>eq', y = NULL, fill = NULL) +
  theme(panel.grid.major.y = element_blank())

p3b <- ggplot(top20, aes(per_capita_t, country, colour = continent)) +
  geom_vline(xintercept = world_per_capita, linetype = 'dotted', colour = 'grey40') +
  geom_segment(aes(x = 0, xend = per_capita_t, yend = country), colour = 'grey75', linewidth = 0.5) +
  geom_point(size = 3) +
  # A borderless white box behind each value keeps the dotted average line from
  # striking through the numbers that sit beside it
  geom_label(aes(label = fmt(per_capita_t, 1)), hjust = -0.3, size = 3.5, colour = 'grey20',
             fill = 'white', border.colour = NA, label.padding = unit(0.1, 'lines')) +
  scale_colour_manual(values = continent_colours, guide = 'none') +
  scale_x_continuous(expand = expansion(mult = c(0, 0.28))) +
  labs(title = 'b. Per capita', x = 't CO<sub>2</sub>eq per person', y = NULL) +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())

p3a + p3b +
  plot_layout(widths = c(2.4, 1), guides = 'collect') +
  plot_annotation(
    caption = paste0(edgar_caption, ' Units: million tonnes of CO<sub>2</sub> equivalent (panel a)',
                     ' and tonnes per person (panel b), both for 2024.<br>',
                     'The dotted line in panel b is the population-weighted world average, ',
                     fmt(world_per_capita, 1), ' t per person.'),
    theme = caption_theme
  ) &
  theme(legend.position = 'bottom')
```

China alone emitted `r pct(top_value('CHN', 'share'))` of the world total in 2024, the United States `r pct(top_value('USA', 'share'))` and India `r pct(top_value('IND', 'share'))`, and the twenty countries together account for `r pct(sum(top20$share), 0)`. Per-capita emissions reorder the list. Saudi Arabia, Australia and Canada, ranked `r ordinal(which(top20$iso3 == 'SAU'))`, `r ordinal(which(top20$iso3 == 'AUS'))` and `r ordinal(which(top20$iso3 == 'CAN'))` by total, emit `r fmt(top_value('SAU', 'per_capita_t'), 1)`, `r fmt(top_value('AUS', 'per_capita_t'), 1)` and `r fmt(top_value('CAN', 'per_capita_t'), 1)` t per person, more than the United States at `r fmt(top_value('USA', 'per_capita_t'), 1)` t and China at `r fmt(top_value('CHN', 'per_capita_t'), 1)` t. India and Pakistan sit at `r fmt(top_value('IND', 'per_capita_t'), 1)` and `r fmt(top_value('PAK', 'per_capita_t'), 1)` t. The population-weighted world average is `r fmt(world_per_capita, 1)` t per person, and `r sum(top20$per_capita_t > world_per_capita)` of the twenty countries lie above it.


### What the largest emitters emit

Figure 4 breaks the 2024 emissions of the eight largest emitters and of the world down by EDGAR sector, summed over the four substances. The composition explains part of the per-capita differences above. Where power generation and industry dominate, emissions follow the fuel mix of the grid, whereas a large agricultural share reflects livestock and rice paddies, which no change of energy source removes.

```{r fig4-sectors, fig.height = 4.8, fig.alt = 'Stacked horizontal bars showing the percentage of 2024 emissions in each of eight EDGAR sectors for the world and the eight largest emitters. Power industry is the largest sector for the world, China, India, Japan, Russia and Indonesia. Transport is the largest for the United States, agriculture makes up about half of the emissions of Brazil, and fuel exploitation is the largest sector in Iran.'}
sector_order <- sectors_2024 %>%
  filter(iso3 == 'GLOBAL TOTAL') %>%
  arrange(desc(mt)) %>%
  pull(sector)

sector_colours <- setNames(c('#0072B2', '#E69F00', '#D55E00', '#009E73',
                             '#8C5BA8', '#CC79A7', '#56B4E9', '#A08C00'), sector_order)

# White text is unreadable on the lighter fills, so each segment label takes
# whichever of white or near-black has the higher WCAG contrast with its fill.
# The two contrasts are equal at a relative luminance of about 0.18.
relative_luminance <- function(hex) {
  channel <- grDevices::col2rgb(hex) / 255
  channel <- ifelse(channel <= 0.04045, channel / 12.92, ((channel + 0.055) / 1.055)^2.4)
  colSums(channel * c(0.2126, 0.7152, 0.0722))
}
sector_text_colours <- setNames(ifelse(relative_luminance(sector_colours) > 0.18, 'grey10', 'white'),
                                names(sector_colours))

top8_iso3 <- as.character(top20$iso3[1:8])

sector_shares <- sectors_2024 %>%
  filter(iso3 %in% c(top8_iso3, 'GLOBAL TOTAL')) %>%
  group_by(iso3) %>%
  mutate(share = 100 * mt / sum(mt)) %>%
  ungroup() %>%
  mutate(country = ifelse(iso3 == 'GLOBAL TOTAL', 'World', country),
         country = factor(country, levels = rev(c('World', as.character(top20$country[1:8])))),
         sector = factor(sector, levels = sector_order))

sector_share <- function(code, sec) sector_shares$share[sector_shares$iso3 == code & sector_shares$sector == sec]

largest_sector <- sector_shares %>%
  filter(iso3 != 'GLOBAL TOTAL') %>%
  group_by(country) %>%
  slice_max(share, n = 1) %>%
  ungroup()
# Country names that take the definite article in running prose.
article <- function(x) ifelse(grepl('^(United|Netherlands|Philippines|Gambia)', x), paste('the', x), x)

# The emitters whose largest sector is `sec`, and the share it takes in each,
# so that the name and the percentage in the prose always come from one lookup.
led_by <- function(sec) filter(largest_sector, sector == sec) %>% arrange(desc(share))
countries_led_by <- function(sec)
  paste(article(as.character(led_by(sec)$country)), collapse = ' and ')
share_led_by <- function(sec, digits = 0) paste(pct(led_by(sec)$share, digits), collapse = ' and ')

# Small counts read better as words in prose.
spell <- function(n) {
  words <- c('one', 'two', 'three', 'four', 'five', 'six', 'seven', 'eight', 'nine', 'ten')
  ifelse(n >= 1 & n <= 10, words[pmax(n, 1)], as.character(n))
}

ggplot(sector_shares, aes(share, country, fill = sector)) +
  geom_col(width = 0.72, colour = 'white', linewidth = 0.5, position = position_stack(reverse = TRUE)) +
  # The label layer keeps every row and blanks the small ones, because
  # position_stack() restacks whatever data it is given: filtering the rows out
  # here instead would place each remaining label as though the omitted sectors
  # did not exist.
  geom_text(aes(label = ifelse(share >= 8, round(share), ''), colour = sector),
            position = position_stack(vjust = 0.5, reverse = TRUE), size = 3.5,
            show.legend = FALSE) +
  scale_fill_manual(values = sector_colours) +
  scale_colour_manual(values = sector_text_colours, guide = 'none') +
  scale_x_continuous(labels = function(x) paste0(x, '%'), expand = expansion(mult = c(0, 0.02))) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = NULL, y = NULL, fill = NULL,
       caption = paste0(edgar_caption, ' Sheet GHG_by_sector_and_country, 2024, summed over substances.<br>',
                        'Units: per cent of each emitter\'s total. Labels are shown for shares of 8% or more.')) +
  theme(panel.grid.major.y = element_blank(), legend.position = 'bottom',
        plot.margin = margin(5.5, 14, 5.5, 5.5))
```

Power generation is the largest sector for the world as a whole, at `r pct(sector_share('GLOBAL TOTAL', 'Power Industry'), 0)` of emissions, and for `r spell(sum(largest_sector$sector == 'Power Industry'))` of the eight countries, with the largest shares in China (`r pct(sector_share('CHN', 'Power Industry'), 0)`) and Japan (`r pct(sector_share('JPN', 'Power Industry'), 0)`). Transport comes first in `r countries_led_by('Transport')` only, at `r share_led_by('Transport')`, and agriculture in `r countries_led_by('Agriculture')` only, at `r share_led_by('Agriculture')`. Fuel exploitation, the extraction and processing of coal, oil and gas, leads in `r countries_led_by('Fuel Exploitation')` only, at `r share_led_by('Fuel Exploitation')`. That last sector also accounts for `r pct(sector_share('IDN', 'Fuel Exploitation'), 0)` of Indonesia's emissions and `r pct(sector_share('RUS', 'Fuel Exploitation'), 0)` of Russia's, against `r pct(sector_share('GLOBAL TOTAL', 'Fuel Exploitation'), 0)` for the world. These compositions matter for the comparison that follows, because methane from livestock and fugitive emissions from fuel extraction do not scale with a country's development in the way that power and transport do.


### Emissions against human development

Per-capita emissions rise with development, and the model in this section measures how far each country departs from the level that its development would predict. Figure 5 plots per-capita emissions on a logarithmic axis against the HDI for 2023. The emissions are averaged over 2015 to 2024 to smooth out the pandemic dip and other year-to-year variation, and each marker is scaled to the country's 2024 population and coloured by continent. The chunk below fits the linear model `log10(per_capita) ~ hdi`, whose line is drawn through the points, and stores each country's residual, the vertical distance from the line in log10 units, for the map in figure 6 and the table after it.

```{r hdi-model}
hdi_data <- countries %>%
  filter(!iso3 %in% c('AIR', 'SEA'), year >= 2015) %>%
  group_by(iso3, country, continent) %>%
  summarise(per_capita_t = mean(per_capita_t), .groups = 'drop') %>%
  left_join(countries %>% filter(year == 2024) %>% select(iso3, total_mt, population),
            by = 'iso3') %>%
  # Jointly reported territories take the HDI of the member whose code they carry
  mutate(hdi_iso3 = ifelse(iso3 == 'SCG', 'SRB', iso3)) %>%
  left_join(hdi %>% select(hdi_iso3 = iso3, hdi_2023, hdicode), by = 'hdi_iso3')

n_no_hdi <- sum(is.na(hdi_data$hdi_2023))
hdi_data <- hdi_data %>% filter(!is.na(hdi_2023))

hdi_model <- lm(log10(per_capita_t) ~ hdi_2023, data = hdi_data)

hdi_data <- hdi_data %>%
  mutate(fitted = fitted(hdi_model),
         residual = residuals(hdi_model),
         expected_t = 10^fitted,
         ratio = 10^residual)

hdi_slope <- coef(hdi_model)[['hdi_2023']]
hdi_r2 <- summary(hdi_model)$r.squared
hdi_n <- nobs(hdi_model)
ratio_of <- function(code) hdi_data$ratio[hdi_data$iso3 == code]
```

Figure 5 is a ggplot converted with `ggplotly`, so hovering over a marker shows the country, its HDI, its per-capita emissions, its 2024 total and its ratio to the fitted line.

```{r fig5-hdi-scatter, fig.height = 6, fig.alt = 'Interactive scatterplot of per-capita greenhouse gas emissions, averaged over 2015 to 2024 on a logarithmic axis, against the Human Development Index for 2023, with one marker per country sized by population and coloured by continent, and a fitted straight line rising from left to right.'}
fit_line <- tibble(hdi_2023 = seq(min(hdi_data$hdi_2023), max(hdi_data$hdi_2023), length.out = 50)) %>%
  mutate(per_capita_t = 10^predict(hdi_model, newdata = .))

hdi_data <- hdi_data %>%
  mutate(tooltip = paste0('<b>', country, '</b>',
                          '<br>HDI 2023: ', fmt(hdi_2023, 3),
                          '<br>Per capita, 2015 to 2024: ', fmt(per_capita_t, 1), ' t',
                          '<br>Total, 2024: ', fmt(total_mt), ' Mt',
                          '<br>', fmt(ratio, 2), ' times the level expected at its HDI'))

p5 <- ggplot(hdi_data, aes(hdi_2023, per_capita_t)) +
  geom_line(data = fit_line, colour = 'grey40', linewidth = 0.8) +
  geom_point(aes(size = population, colour = continent, text = tooltip), alpha = 0.7) +
  scale_y_log10(labels = comma) +
  scale_size_area(max_size = 20, guide = 'none') +
  scale_colour_manual(values = continent_colours, name = NULL) +
  # On two lines, because on one the title is longer than the plot is tall
  labs(x = 'Human Development Index, 2023',
       y = 'Per-capita emissions, 2015 to 2024\n(t CO<sub>2</sub>eq, log scale)') +
  # A complete theme, so that ggplotly does not merge the ggtext elements of the global theme
  theme_minimal(base_size = 13)

fig5 <- ggplotly(p5, tooltip = 'text') %>%
  layout(legend = list(orientation = 'h', y = -0.18, x = 0.5, xanchor = 'center',
                       itemsizing = 'constant'),
         margin = list(l = 80, r = 20, b = 70, t = 20),
         hoverlabel = list(align = 'left')) %>%
  # The full plotly bundle is about 3.4 MB, most of it chart types this figure
  # does not use. partial_bundle() substitutes the smallest build that still
  # covers the traces present. It fetches that build from the plotly CDN, so it
  # falls back to the full bundle when the render has no network access.
  lighter_bundle()

# knitr would otherwise fix the widget at fig.width x dpi, 768 px, which is
# wider than the 720 px text column and runs off the side of a phone screen.
# At 100% the widget follows its column and plotly redraws to that width.
fig5$width <- '100%'

# At the size that suits the desktop column, the x-axis title is wider than the
# plot on a 360 px phone and loses its last characters, so narrow screens get
# smaller axis titles
htmlwidgets::onRender(fig5, 'function(el) {
  if (el.offsetWidth < 400) {
    Plotly.relayout(el, {"xaxis.title.font.size": 15, "yaxis.title.font.size": 15});
  }
}')
```

<p class="caption">Figure 5. Source: EDGAR 2025 (Crippa et al., 2025) and the Human Development Report 2025 (United Nations Development Programme, 2025). Units: tonnes of CO<sub>2</sub> equivalent per person per year, averaged over 2015 to 2024. Marker area proportional to 2024 population.</p>

The model is fitted to the `r hdi_n` countries with both an HDI and emissions. The `r n_no_hdi` territories without an HDI are left out: mostly dependencies and overseas territories such as Puerto Rico, Greenland and Macao, plus Taiwan, North Korea and Western Sahara. The slope is `r fmt(hdi_slope, 2)` log10 units per unit of HDI, so each additional 0.1 of HDI multiplies expected per-capita emissions by `r fmt(10^(hdi_slope / 10), 2)`. The model explains `r pct(100 * hdi_r2, 0)` of the variance in log emissions (*R*<sup>2</sup> = `r fmt(hdi_r2, 2)`, *n* = `r hdi_n`). Of the countries, `r pct(100 * mean(abs(hdi_data$residual) < log10(2)), 0)` lie within a factor of two of the line. Qatar and Kuwait emit `r fmt(ratio_of('QAT'), 1)` and `r fmt(ratio_of('KWT'), 1)` times the level expected at their HDI, Russia `r fmt(ratio_of('RUS'), 1)`, and the United States `r fmt(ratio_of('USA'), 1)`. Sweden and Switzerland, by comparison, emit `r fmt(ratio_of('SWE'), 2)` and `r fmt(ratio_of('CHE'), 2)` times theirs. China sits at `r fmt(ratio_of('CHN'), 1)` and India at `r fmt(ratio_of('IND'), 2)`.

The residual describes how a country compares with others at the same HDI, and four caveats apply to any reading of it. The HDI includes gross national income, so it partly encodes the energy use that it is meant to contextualise, and the line therefore absorbs some of the effect of wealth that a reader might want to see. EDGAR is production-based, so the emissions embodied in goods that a country imports are attributed to the exporter, which favours importers of manufactured goods in this comparison and counts against exporters such as China. The exclusion of land use, land-use change and forestry matters for countries such as Brazil and Indonesia. Finally, the model is a single straight line through `r hdi_n` points. Oil and gas exporters, pastoral economies, and small states whose per-capita figure depends on one plant or fleet all pull on it in ways that a fuller treatment would handle differently.


### Mapping the residuals

Figure 6 maps the residuals of the model on a diverging scale centred on zero, red where a country emits more than expected at its HDI and blue where it emits less. Hovering shows the residual in log10 units and the ratio 10<sup>residual</sup>, which is easier to read. Countries without an HDI or without emissions data are shown in light grey. The members of a territory that EDGAR reports jointly carry the value of that territory, and their hover text says so.

```{r fig6-map, fig.alt = 'Interactive world map in which each country is coloured by its residual from the model of per-capita emissions against the Human Development Index, red for higher than expected and blue for lower than expected, with countries lacking data in light grey.'}
# Natural Earth's medium geometry is about 3 MB of coordinates in the widget
# below. Simplifying it would cut that substantially, but every tolerance-based
# method available here damages the map: whole-feature st_simplify() silently
# drops 123 of the 1,620 polygon parts, among them 16 of Kiribati's 19 islands,
# and the per-part variants that keep them leave invalid geometries behind. A
# topology-aware simplifier such as rmapshaper::ms_simplify(keep_shapes = TRUE)
# is the tool for this; until the geometry can be checked under one, the map
# ships at full fidelity.
world_sf <- ne_countries(scale = 'medium', returnclass = 'sf') %>%
  # France and Norway carry -99 as iso_a3 in Natural Earth; iso_a3_eh holds the code
  mutate(iso_a3 = ifelse(iso_a3 == '-99', iso_a3_eh, iso_a3)) %>%
  filter(iso_a3 != 'ATA') %>%
  select(iso_a3, name_long) %>%
  mutate(edgar_iso3 = coalesce(combined$iso3[match(iso_a3, combined$member_iso3)], iso_a3),
         jointly = iso_a3 %in% combined$member_iso3) %>%
  left_join(hdi_data %>% select(iso3, country, hdi_2023, per_capita_t, expected_t, residual, ratio),
            by = c('edgar_iso3' = 'iso3'))

residual_limit <- max(abs(world_sf$residual), na.rm = TRUE)
residual_palette <- colorNumeric(rev(brewer.pal(11, 'RdBu')),
                                 domain = c(-residual_limit, residual_limit),
                                 na.color = '#d9d9d9')

map_labels <- with(world_sf, ifelse(
  is.na(residual),
  paste0('<b>', name_long, '</b><br>no data'),
  paste0('<b>', name_long, '</b>',
         ifelse(jointly, paste0('<br>Reported by EDGAR as ', country), ''),
         '<br>HDI 2023: ', fmt(hdi_2023, 3),
         '<br>Per capita, 2015 to 2024: ', fmt(per_capita_t, 1), ' t (expected ', fmt(expected_t, 1), ' t)',
         '<br>Residual: ', formatC(residual, format = 'f', digits = 2, flag = '+'), ' log10 units',
         '<br>', fmt(ratio, 2), ' times the level expected at its HDI')
# unname(): with() carries the sf row names onto the result, and htmlwidgets
# rejects a partially named list when it scans the widget for JS() markers
)) %>% unname() %>% lapply(HTML)

# A fixed zoom shows a 512 px wide world, which a phone crops on both sides, so
# the map instead fits the inhabited latitudes to its container at a fractional
# zoom. On a narrow screen the fitted world is only about 240 px tall, so the
# container is shorter there and the world is fitted above the legend, which
# would otherwise cover South America.
leaflet(world_sf, width = '100%', height = 520, elementId = 'residual-map',
        options = leafletOptions(minZoom = 0, maxZoom = 6, zoomSnap = 0.1, zoomDelta = 0.5,
                                 worldCopyJump = TRUE)) %>%
  addPolygons(fillColor = ~ residual_palette(residual), fillOpacity = 0.9,
              color = 'white', weight = 0.6, smoothFactor = 0.3,
              label = map_labels,
              highlightOptions = highlightOptions(weight = 2, color = '#333333', bringToFront = TRUE)) %>%
  # One legend box, with the no-data swatch under the scale, covers less of the map
  addLegend(pal = residual_palette, values = c(-residual_limit, residual_limit, NA),
            na.label = 'no data', title = 'Residual (log10)', position = 'bottomleft',
            opacity = 0.9) %>%
  fitBounds(lng1 = -170, lat1 = -56, lng2 = 180, lat2 = 84) %>%
  htmlwidgets::onRender(
    'function(el, x) {
       if (window.matchMedia("(max-width: 600px)").matches) {
         this.fitBounds([[-56, -170], [84, 180]], {paddingBottomRight: [0, 150], animate: false});
       }
     }') %>%
  # leaflet floats the colour scale and its labels without clearing the float,
  # so the no-data row would otherwise sit beside the scale instead of under it.
  # The site's dark theme gives every link near-white text, which would leave the
  # white zoom buttons and the attribution blank, so the map's links stay dark.
  htmlwidgets::prependContent(tags$style(HTML(
    '.leaflet-container { background: #f4f6f8; }
     #residual-map .info.legend > div:last-child { clear: both; padding-top: 4px; }
     #residual-map .leaflet-control a { color: #333 !important; -webkit-text-fill-color: #333 !important;
       text-decoration: none !important; }
     @media (max-width: 600px) { #residual-map { height: 410px !important; } }')))
```

<p class="caption">Figure 6. Source: EDGAR 2025 (Crippa et al., 2025), the Human Development Report 2025 (United Nations Development Programme, 2025) and Natural Earth boundaries via rnaturalearth. Units: residual of log10 per-capita emissions (2015 to 2024 mean) from the linear model on HDI 2023. The ratio of actual to expected emissions is 10<sup>residual</sup>. Boundaries carry no political meaning.</p>

The largest positive residual belongs to `r hdi_data$country[which.max(hdi_data$residual)]`, at `r fmt(max(hdi_data$ratio), 1)` times its expected level, and the largest negative one to `r hdi_data$country[which.min(hdi_data$residual)]`, at `r fmt(min(hdi_data$ratio), 2)` times. The table below lists the ten countries at each end. Oil and gas exporters fill most of the positive end, together with small states whose per-capita figure is set by one power plant or one fleet, such as Palau, with `r fmt(hdi_data$population[hdi_data$iso3 == 'PLW'] / 1e3)` thousand inhabitants. Two pastoral economies also appear there: in Mongolia, fuel extraction and livestock account for `r pct(sum(sectors_2024$mt[sectors_2024$iso3 == 'MNG' & sectors_2024$sector %in% c('Fuel Exploitation', 'Agriculture')]) / sum(sectors_2024$mt[sectors_2024$iso3 == 'MNG']) * 100, 0)` of emissions, and in Chad, agriculture accounts for `r pct(sectors_2024$mt[sectors_2024$iso3 == 'TCD' & sectors_2024$sector == 'Agriculture'] / sum(sectors_2024$mt[sectors_2024$iso3 == 'TCD']) * 100, 0)`. Methane from herds enters the total in full, and the model expects little of it at a low HDI, so a country whose emissions are mostly livestock lands above the line. The negative end mixes small island states with countries of South Asia and Sub-Saharan Africa whose emissions are low for their HDI. At the top of the HDI range, it also holds Switzerland, whose electricity comes largely from hydropower and nuclear plants and which EDGAR reports together with Liechtenstein. Because EDGAR reports Sudan and South Sudan as one territory, their combined per-capita figure is compared with the HDI of Sudan and mapped onto both countries.

```{r residual-table}
bind_rows(hdi_data %>% slice_max(residual, n = 10),
          hdi_data %>% slice_min(residual, n = 10) %>% arrange(desc(residual))) %>%
  transmute(Country = country, Continent = continent, `HDI 2023` = fmt(hdi_2023, 3),
            `Per capita (t)` = fmt(per_capita_t, 2), `Expected (t)` = fmt(expected_t, 2),
            Ratio = fmt(ratio, 2)) %>%
  kable(align = 'llrrrr',
        caption = paste('The ten countries emitting most above (top) and most below (bottom)',
                        'the level expected at their HDI. Per capita is the 2015 to 2024 mean',
                        'in t CO2eq per person, expected is the fitted value and ratio is',
                        'actual over expected.'))
```

Whereas a few recurring causes account for the positive end of the table, the negative end has no single cause and spans HDI values from `r fmt(min(hdi_data$hdi_2023[order(hdi_data$residual)][1:10]), 2)` to `r fmt(max(hdi_data$hdi_2023[order(hdi_data$residual)][1:10]), 2)`.


## Discussion

Read together, the figures give a contextual account of emissions. Total emissions describe the size of each country's annual contribution to the global total, and per-capita values describe the average emissions of its residents. Sector shares suggest where mitigation could begin, and the HDI residuals show departures from a broad relationship between development and emissions. These are complementary summaries, and their interpretation depends on the boundary choices documented above, especially the exclusion of land use, land-use change and forestry, and the treatment of international transport.

### How the figures were produced

```{r session, include = FALSE}
packages_used <- c('dplyr', 'tidyr', 'ggplot2', 'patchwork', 'scales', 'ggrepel',
                   'ggtext', 'plotly', 'htmltools', 'leaflet', 'sf', 'rnaturalearth',
                   'rnaturalearthdata', 'RColorBrewer', 'knitr', 'rmarkdown')
package_versions <- paste(packages_used, sapply(packages_used, function(p) as.character(packageVersion(p))))
package_text <- paste0(paste(head(package_versions, -1), collapse = ', '), ' and ', tail(package_versions, 1))
```

The EDGAR sheets store years as columns, so each was reshaped to one row per territory and year before the totals and per-capita figures were joined by country code and year. Population-weighted means and population shares were computed from the derived population, and the world population is the sum over the territories that report a per-capita figure. Continents, income groups and the HDI were joined by ISO 3166 alpha-3 code. A few exceptions were handled by hand. Taiwan has no region in the ISO 3166 dataset. EDGAR has no row for Kosovo and files Curaçao under the code of the former Netherlands Antilles. The jointly reported territories keep the code under which EDGAR reports them. The model of figure 5 is an ordinary least-squares regression of log10 per-capita emissions, averaged over 2015 to 2024, on the HDI for 2023, fitted to `r hdi_n` countries. The document was rendered with `r R.version.string` and the packages `r package_text`.


## References

Crippa, M., Guizzardi, D., Pagani, F., Banja, M., Muntean, M., Schaaf, E., Quadrelli, R., Risquez Martin, A., Taghavi-Moharamli, P., Köykkä, J., Grassi, G., Rossi, S., Melo, J., Oom, D., Branco, A., Suárez-Moreno, M., Sedano, F., San-Miguel, J., Manca, G., … Pekar, F. (2025). *GHG emissions of all world countries: 2025 report*. Publications Office of the European Union. https://doi.org/10.2760/9816914

Duncalfe, L. (n.d.). *ISO-3166-Countries-with-Regional-Codes* [Data set]. GitHub. https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes

United Nations Development Programme. (2025). *Human development report 2025: A matter of choice: People and possibilities in the age of AI*. United Nations. https://doi.org/10.18356/9789211542639

World Bank. (2025). *World Bank country and lending groups: Current classification by income* (Version 2025-07-02) [Data set]. https://datacatalogfiles.worldbank.org/ddh-published/0037712/DR0095333/CLASS_2025_07_02.xlsx
