---
title: 'Who is through with convergence warnings? Publications that used allFit to compare optimisers'
author: ''
date: '2026-09-02'
slug: publications-that-used-allfit-to-compare-optimisers
categories:
  - statistics
  - bibliometrics
tags:
  - lme4
  - convergence
  - linear mixed-effects models
  - research methods
  - bibliometrics
subtitle: ''
summary: 'A convergence warning from lme4 is a symptom, and refitting the model with every available optimiser through allFit is the usual way to find out whether it matters. Reviewers sometimes ask for a precedent. This post keeps a record of published studies that used allFit, gathered from Scopus, Europe PMC and OpenAlex by a workflow that can be rerun, and explains what each source can and cannot find.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 95)
xaringanExtra::use_clipboard()
```

Mixed-effects models with the random-effects structure that the design justifies (Barr et al., 2013) often end their fit with a warning from lme4 that the optimiser may not have converged. The warning reports that the gradient at the solution exceeded a tolerance, which does not by itself mean that the fit is wrong. The lme4 check produces false positives, as the help page on convergence explains, especially with large data sets and many random effects. A model that has struggled to find enough information to estimate every variance and correlation has other tell-tales, such as correlations of exactly 1 or variances of exactly 0 (Bates et al., 2015; Matuschek et al., 2017), and the usual advice is to look at those before simplifying the model (Brauer & Curtin, 2018; Singmann & Kellen, 2019).

The most direct check is to refit the model with every optimiser that lme4 can call, through `allFit()`, and to compare the fixed-effect estimates across the fits. If the optimisers agree to the precision that matters for the conclusions, the warning can be reported and set aside, whereas a disagreement between them points to a problem that no change of optimiser will solve. I wrote about the practice and published a [plotting function for the comparison](/2023/a-new-function-to-plot-convergence-diagnostics-from-lme4-allfit/) some years ago, and since then I have been asked several times, by co-authors and by reviewers, whether other published studies have proceeded in the same way. This post keeps a list of the ones that can be found by searching, so that the next person asked can point to it.

## Where a function name can be found

A function name is a poor search term for bibliographic databases, because it appears in the methods section of an article and almost never in its title, abstract or keywords. Three sources with different coverage are therefore searched, by a script in the [website's repository](https://github.com/pablobernabeu/website-files/blob/master/scripts/update_post_searches.R) that a GitHub Actions workflow runs on request.

Scopus, searched through the [scopusflow package](/2026/scopusflow-a-literature-search-you-can-rerun/), indexes titles, abstracts, keywords, references and a few other fields, but not the body of the article, so a search for the function name in any indexed field finds only articles that mention it in one of those. A second Scopus query, drafted for the first version of this post in 2023, is a proxy: it looks for abstracts that mention lme4, lmerTest or brms alongside maximal random slopes and convergence, on the grounds that an abstract discussing convergence at that level of detail is likely to belong to a study that dealt with it. Europe PMC indexes the full text of a large body of life-science articles, most of them open access but not all, and its search interface accepts a full-text query. OpenAlex offers a full-text search over the articles for which it holds the text. The Europe PMC query pairs the function name with a mixed-model term, which removes some unrelated uses of the string. The OpenAlex query carries no such term, so its results are matches on the string alone.

```{r, eval = FALSE}
library(scopusflow)   # reads the key from SCOPUS_API_KEY
scopus_any_field <- scopus_fetch('ALL("allFit")')
scopus_proxy <- scopus_fetch(
  paste('("lme4" OR "lmerTest" OR "brms") AND maximal AND "random slopes"',
        'AND (convergence OR converge OR converged OR converging)'),
  field = 'TITLE-ABS-KEY')
```

```{r, eval = FALSE}
# Europe PMC and OpenAlex, through their public REST interfaces
europepmc_query <- paste(
  '"allFit" AND (lme4 OR "mixed-effects" OR "mixed effects" OR',
  '"mixed model" OR "mixed models" OR "multilevel")')
openalex_filter <- 'fulltext.search:"allFit"'
```

## The list

The workflow commits its results to this post's directory, and the tables below are built from those files, so the record can be refreshed without editing the post.

```{r}
library(dplyr)
library(ggplot2)

read_if_present <- function(file, source) {
  path <- file.path('searches', file)
  if (!file.exists(path)) return(NULL)
  x <- read.csv(path, stringsAsFactors = FALSE)
  x$source <- source
  x
}
retrieved <- readLines('searches/retrieved.txt')

scopus_any   <- read_if_present('scopus_allfit_any_field.csv', 'Scopus, any indexed field')
scopus_proxy <- read_if_present('scopus_convergence_proxy.csv', 'Scopus, convergence proxy')
europepmc    <- read_if_present('europepmc_allfit_fulltext.csv', 'Europe PMC, full text')
openalex     <- read_if_present('openalex_allfit_fulltext.csv', 'OpenAlex, full text')

standardise <- function(x) {
  if (is.null(x)) return(NULL)
  tibble(source = x$source,
         doi = tolower(sub('^https?://doi.org/', '', x$doi)),
         year = as.integer(x$year),
         authors = x$authors,
         title = x$title,
         venue = if ('journal' %in% names(x)) x$journal else x$publication)
}
records <- bind_rows(lapply(list(scopus_any, scopus_proxy, europepmc, openalex),
                            standardise))
records |> count(source, name = 'records')

# Do any of the Scopus any-field titles mention a mixed model at all?
mixed_terms <- 'mixed|lme4|multilevel|random effect|convergence|optimi[sz]'
sum(grepl(mixed_terms, scopus_any$title, ignore.case = TRUE))
head(sort(table(scopus_any$publication), decreasing = TRUE), 4)
```

The three sources were queried on `r format(as.Date(substr(retrieved, 1, 10)), '%e %B %Y') |> trimws()`. The Scopus search for the function name in any indexed field illustrates why a bare string is a poor query: it returned `r if (is.null(scopus_any)) 0 else nrow(scopus_any)` records, the largest group of them in pharmacology and medicinal chemistry, and not one of their titles mentions a mixed model. Scopus appears to match the string loosely against unrelated text, so those records are left out of the list below. The proxy query returned `r if (is.null(scopus_proxy)) 'no records' else paste(nrow(scopus_proxy), if (nrow(scopus_proxy) == 1) 'record' else 'records')`, which shows how rarely an abstract describes the treatment of convergence at that level of detail. The two full-text sources are the ones that come closest to answering the question, since they reach the methods section, where the function name is actually written. Their results are merged on the DOI and screened once, by dropping any record published before the function reached lme4, since a match in such an article is something other than a use of it.

```{r}
# allFit reached lme4 in version 1.1-7, released in July 2014, so a record
# published before then cannot be a use of it, whatever matched the string.
allfit_year <- 2014L

used_allfit <- records |>
  filter(source %in% c('Europe PMC, full text', 'OpenAlex, full text')) |>
  mutate(key = ifelse(is.na(doi) | doi == '', paste(source, title), doi)) |>
  group_by(key) |>
  summarise(year = first(na.omit(year)), authors = first(authors),
            title = first(title), venue = first(na.omit(venue)),
            doi = first(doi), sources = paste(sort(unique(source)), collapse = '; '),
            .groups = 'drop') |>
  arrange(desc(year), authors)

too_early <- used_allfit |> filter(!is.na(year), year < allfit_year)
used_allfit <- used_allfit |> filter(is.na(year) | year >= allfit_year)
c(matched = nrow(used_allfit) + nrow(too_early),
  before_allfit_existed = nrow(too_early), kept = nrow(used_allfit))
table(used_allfit$sources)
```

```{r, fig.width = 7, fig.height = 3.4, fig.alt = paste0('Bar chart of the number of publications per year, from ', min(used_allfit$year, na.rm = TRUE), ' to ', max(used_allfit$year, na.rm = TRUE), ', whose full text mentions allFit, according to Europe PMC and OpenAlex')}
used_allfit |>
  filter(!is.na(year)) |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_col(fill = '#0072B2') +
  scale_x_continuous(breaks = scales::breaks_pretty(8)) +
  labs(x = NULL, y = 'Publications mentioning allFit') +
  theme_minimal(base_size = 13)
```

The table lists the publications kept by the full-text searches, most recent first, with the sources in which each was found. Authors are given as the sources returned them. The two sources overlap in a minority of records, which says as much about their coverage as about the literature.

```{r}
short_authors <- function(a) {
  a <- ifelse(is.na(a), '', a)
  first <- sub('[;,].*$', '', a)
  ifelse(grepl('[;,]', a), paste(first, 'et al.'), first)
}
used_allfit |>
  transmute(Year = year,
            Authors = short_authors(authors),
            Title = title,
            Venue = venue,
            DOI = ifelse(is.na(doi) | doi == '', '',
                         sprintf('[%s](https://doi.org/%s)', doi, doi)),
            Sources = sources) |>
  knitr::kable()
```

The proxy query is shown for completeness. Its single hit is an erratum notice that carries the original article's abstract, which is a fair indication of how rarely an abstract describes the treatment of convergence in this much detail. A wider proxy would be a reasonable place to look for further precedents, since studies that compared optimisers rarely name the function in a field that Scopus indexes.

```{r}
records |>
  filter(source == 'Scopus, convergence proxy') |>
  transmute(Year = year, Authors = short_authors(authors), Title = title, Venue = venue,
            DOI = ifelse(is.na(doi) | doi == '', '',
                         sprintf('[%s](https://doi.org/%s)', doi, doi))) |>
  knitr::kable()
```

## What the list misses

The count is a lower bound. Europe PMC holds full text mostly for open-access articles, and mostly in journals within its scope, so a study behind a paywall in a linguistics or psychology journal is usually invisible to it unless a preprint or an accepted manuscript was deposited. OpenAlex holds text for a subset of the literature that it does not fully document. Scopus cannot see the methods section at all, and a full-text match can be an article that uses the function in a context other than a convergence check, or that mentions it in a footnote. A study that compared optimisers with its own loop, or with the `optimx` package directly, is not found by any of the three, and neither is a study that reported the comparison only in supplementary material. Records published before the function existed have been dropped, but the table has not otherwise been screened by hand, so it lists the publications whose full text matches the name rather than the publications that are known to have used it. The list is therefore a starting point for the reader who needs a precedent, and it can be refreshed by dispatching the workflow again.

### References

Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language*, *68*(3), 255–278. https://doi.org/10.1016/j.jml.2012.11.001

Bates, D., Kliegl, R., Vasishth, S., & Baayen, H. (2015). Parsimonious mixed models. *arXiv*. https://doi.org/10.48550/arXiv.1506.04967

Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, *67*(1), 1–48. https://doi.org/10.18637/jss.v067.i01

Brauer, M., & Curtin, J. J. (2018). Linear mixed-effects models and the analysis of nonindependent data: A unified framework to analyze categorical and continuous independent variables that vary within-subjects and/or within-items. *Psychological Methods*, *23*(3), 389–411. https://doi.org/10.1037/met0000159

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language*, *94*, 305–315. https://doi.org/10.1016/j.jml.2017.01.001

Singmann, H., & Kellen, D. (2019). An introduction to mixed models for experimental psychology. In D. H. Spieler & E. Schumacher (Eds.), *New methods in cognitive psychology* (pp. 4–31). Routledge. https://doi.org/10.4324/9780429318405-2
