---
title: 'Done with convergence warnings? Publications that used allFit to compare optimisers'
author: ''
date: '2026-09-02'
slug: publications-that-used-allfit-to-compare-optimisers
code_folding: hide
categories:
  - statistics
  - bibliometrics
tags:
  - s
  - lme4
  - convergence
  - linear mixed-effects models
  - research methods
  - bibliometrics
subtitle: ''
summary: 'A convergence warning from lme4 does not by itself show that a model is wrong. One check that the lme4 authors recommend is to refit the model with every available optimiser through allFit and compare the estimates that matter. This post records the candidate publications found by a reproducible search of Scopus, Europe PMC and OpenAlex, and explains why the list can only be a starting point for screening.'
abstract: 'This reproducible bibliographic audit searches Scopus, Europe PMC and OpenAlex for publications whose indexed records or full text mention lme4::allFit. It separates full-text matches from those of a Scopus proxy query, deduplicates records by DOI and presents the result as a lower-bound screening set, since a listed study mentions allFit without necessarily having used it to resolve a convergence warning.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 95)
xaringanExtra::use_clipboard()
```

## Introduction

Mixed-effects models with the random-effects structure that the design justifies (Barr et al., 2013) can end their fit with a warning from lme4 (Bates, Mächler, et al., 2015) that the optimiser may not have converged. The warning reports that a numerical check exceeded a tolerance, but that alone does not establish that the fitted model is wrong. According to the lme4 documentation, the check relies on approximate derivatives that are sometimes unreliable, and it can flag apparently well-behaved fits to large data sets. Even so, a warning calls for investigation before it is set aside. A model with too little information to estimate every variance and correlation may also show boundary estimates, such as a variance near zero or a correlation near one (Bates, Kliegl, et al., 2015; Matuschek et al., 2017; Singmann & Kellen, 2019).

Methods tutorials describe several corrective steps for convergence problems (Brauer & Curtin, 2018; Singmann & Kellen, 2019). The lme4 documentation recommends a similar series of steps. Its authors describe one of them as the gold standard, namely refitting the model with every available optimiser through `allFit()` and comparing the estimates that matter for the conclusions. Practical agreement between the optimisers is consistent with a false-positive warning, although it does not show that the model is otherwise appropriate. Disagreement calls for a closer look at the data, the model specification and the optimisation settings. Some years ago, I wrote about the practice and published a [plotting function for the comparison](/2023/a-new-function-to-plot-convergence-diagnostics-from-lme4-allfit/). This post keeps a reproducible record of the candidate publications that a search can find, so that readers can inspect the evidence for themselves.

## Development: finding a function name

A function name is a poor search term for bibliographic databases, because it appears in the methods section of an article and almost never in its title, abstract or keywords. Three sources with different coverage were therefore searched with a [script in the website's repository](https://github.com/pablobernabeu/website-files/blob/master/scripts/update_post_searches.R).

Scopus, which I queried through my [scopusflow](/2026/scopusflow-a-literature-search-you-can-rerun/) package (Bernabeu, 2026), indexes titles, abstracts, keywords, references and a few other fields but not the body of an article. A search for the function name in any indexed field can only find the articles that mention it in one of those fields. A second Scopus query serves as a proxy. It looks for titles, abstracts and keywords that mention lme4, lmerTest or brms alongside maximal random slopes and convergence. The assumption is that an abstract discussing convergence in that much detail probably belongs to a study that dealt with it. Europe PMC holds the full text of a large body of life-science articles, most but not all of them open access, and its search accepts full-text queries. OpenAlex offers a full-text search over the articles whose text it holds. Both full-text queries pair the function name with a mixed-model term, which removes some unrelated uses of the string without removing all of them.

```{r, eval = FALSE}
library(scopusflow)   # reads the key from SCOPUS_API_KEY
scopus_any_field <- scopus_fetch('ALL("allFit")')
scopus_proxy <- scopus_fetch(
  paste('("lme4" OR "lmerTest" OR "brms") AND maximal AND "random slopes"',
        'AND (convergence OR converge OR converged OR converging)'),
  field = 'TITLE-ABS-KEY')
```

The two full-text queries are read from the files saved with the results, so they are exactly the queries that ran.

```{r, comment = ''}
cat(readLines('searches/europepmc_query.txt'), sep = '\n')
cat(readLines('searches/openalex_filter.txt'), sep = '\n')
```

## Results: the search record

Each search saved its results to a CSV file, and the tables below are built from those files.

```{r}
library(dplyr)
library(ggplot2)

read_if_present <- function(file, source) {
  path <- file.path('searches', file)
  if (!file.exists(path)) return(NULL)
  x <- read.csv(path, stringsAsFactors = FALSE)
  x$source <- source
  x
}
retrieved <- readLines('searches/retrieved.txt')

scopus_any   <- read_if_present('scopus_allfit_any_field.csv', 'Scopus, any indexed field')
scopus_proxy <- read_if_present('scopus_convergence_proxy.csv', 'Scopus, convergence proxy')
europepmc    <- read_if_present('europepmc_allfit_fulltext.csv', 'Europe PMC, full text')
openalex     <- read_if_present('openalex_allfit_fulltext.csv', 'OpenAlex, full text')

standardise <- function(x) {
  if (is.null(x)) return(NULL)
  tibble(source = x$source,
         doi = tolower(sub('^https?://doi.org/', '', x$doi)),
         year = as.integer(x$year),
         authors = x$authors,
         title = x$title,
         venue = if ('journal' %in% names(x)) x$journal else x$publication)
}
records <- bind_rows(lapply(list(scopus_any, scopus_proxy, europepmc, openalex),
                            standardise))
records |> count(source, name = 'records')

# Count the Scopus any-field titles that mention a mixed model at all.
mixed_terms <- 'mixed|lme4|multilevel|random effect|convergence|optimi[sz]'
sum(grepl(mixed_terms, scopus_any$title, ignore.case = TRUE))
head(sort(table(scopus_any$publication), decreasing = TRUE), 4)
```

The three sources were queried on `r format(as.Date(substr(retrieved, 1, 10)), '%e %B %Y') |> trimws()`. The Scopus search for the function name in any indexed field shows why a bare string makes a poor query. It returned `r if (is.null(scopus_any)) 0 else nrow(scopus_any)` records, the largest group of them in pharmacology and medicinal chemistry, and not one of their titles mentions a mixed model. Scopus appears to match the string loosely against unrelated text, so those records are left out of the list below. The proxy query returned `r if (is.null(scopus_proxy)) 'no records' else paste(nrow(scopus_proxy), if (nrow(scopus_proxy) == 1) 'record' else 'records')`, which shows how rarely an abstract describes the treatment of convergence in that much detail. The two full-text sources come closest to answering the question, because they reach the methods section, where the function name usually appears. Their results are merged on the DOI and screened once, by dropping any record published before allFit existed, as a match in such an article cannot be a use of the function.

```{r}
# allFit appeared in lme4's GitHub repository in April 2014 and first shipped
# with lme4 1.1-8 in June 2015, so a record published before 2014 cannot be a
# use of it, whatever matched the string.
allfit_year <- 2014L

used_allfit <- records |>
  filter(source %in% c('Europe PMC, full text', 'OpenAlex, full text')) |>
  mutate(key = ifelse(is.na(doi) | doi == '', paste(source, title), doi)) |>
  group_by(key) |>
  summarise(year = first(na.omit(year)), authors = first(authors),
            title = first(title), venue = first(na.omit(venue)),
            doi = first(doi), sources = paste(sort(unique(source)), collapse = '; '),
            .groups = 'drop') |>
  arrange(desc(year), authors)

too_early <- used_allfit |> filter(!is.na(year), year < allfit_year)
used_allfit <- used_allfit |> filter(is.na(year) | year >= allfit_year)
c(matched = nrow(used_allfit) + nrow(too_early),
  before_allfit_existed = nrow(too_early), kept = nrow(used_allfit))
table(used_allfit$sources)
```

The figure below shows how many of the kept publications appeared in each year.

```{r, fig.width = 7, fig.height = 3.4, fig.alt = paste0('Bar chart of the number of publications per year, from ', min(used_allfit$year, na.rm = TRUE), ' to ', max(used_allfit$year, na.rm = TRUE), ', whose full text mentions allFit, according to Europe PMC and OpenAlex')}
per_year <- used_allfit |>
  filter(!is.na(year)) |>
  count(year)

# The year of the search was still under way, so its label says where the
# count stops.
searched <- as.Date(substr(retrieved, 1, 10))
year_label <- function(y) {
  ifelse(y == as.integer(format(searched, '%Y')),
         paste0(y, '\nto ', trimws(format(searched, '%e %b'))), y)
}

ggplot(per_year, aes(year, n)) +
  geom_col(fill = '#0072B2', width = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 3.4) +
  scale_x_continuous(breaks = min(per_year$year):max(per_year$year),
                     labels = year_label) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = NULL, title = 'Publications mentioning allFit') +
  theme_minimal(base_size = 12) +
  theme(plot.title.position = 'plot', plot.title = element_text(size = rel(1)),
        axis.text.y = element_blank(), panel.grid = element_blank())
```

The table below lists the same publications, most recent first, with the sources in which each was found. Authors are given as the sources returned them. The two sources overlap in a minority of records, which says as much about their coverage as about the literature.

```{r}
short_authors <- function(a) {
  a <- ifelse(is.na(a), '', a)
  first <- sub('[;,].*$', '', a)
  ifelse(grepl('[;,]', a), paste(first, 'et al.'), first)
}
used_allfit |>
  transmute(Year = year,
            Authors = short_authors(authors),
            Title = title,
            Venue = venue,
            DOI = ifelse(is.na(doi) | doi == '', '',
                         sprintf('[%s](https://doi.org/%s)', doi, doi)),
            Sources = sources) |>
  knitr::kable()
```

The only hit of the proxy query, listed below, is an erratum notice that carries the abstract of the original article. A wider proxy query would be a reasonable place to look for further precedents, since studies that compared optimisers rarely name the function in a field that Scopus indexes.

```{r}
records |>
  filter(source == 'Scopus, convergence proxy') |>
  transmute(Year = year, Authors = short_authors(authors), Title = title, Venue = venue,
            DOI = ifelse(is.na(doi) | doi == '', '',
                         sprintf('[%s](https://doi.org/%s)', doi, doi))) |>
  knitr::kable()
```

## Discussion: what the list misses

The count is a lower bound, because the searches miss part of the literature. Europe PMC holds full text mostly for open-access articles in journals within its scope, so a paywalled study in a linguistics or psychology journal is usually invisible to it unless a preprint or an accepted manuscript was deposited. OpenAlex holds the text of a subset of the literature that it does not fully document, and Scopus cannot see the methods section at all. None of the three can find a study that compared optimisers with its own loop or by calling the `optimx` package directly, or one that reported the comparison only in supplementary material. At the same time, a full-text match can be an article that uses the function for something other than a convergence check, or that mentions it only in a footnote. Records published before the function existed have been dropped, but the table has not otherwise been screened by hand, so some of the publications it lists may mention the function without having used it. The list is therefore a starting point for the reader who needs a precedent.

## References

Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language*, *68*(3), 255–278. https://doi.org/10.1016/j.jml.2012.11.001

Bates, D., Kliegl, R., Vasishth, S., & Baayen, H. (2015). *Parsimonious mixed models* [Preprint]. arXiv. https://doi.org/10.48550/arXiv.1506.04967

Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, *67*(1), 1–48. https://doi.org/10.18637/jss.v067.i01

Bernabeu, P. (2026). *scopusflow: A reproducible workflow layer for Scopus bibliographic searches* (Version 0.4.0) [Computer software]. CRAN. https://doi.org/10.32614/CRAN.package.scopusflow

Brauer, M., & Curtin, J. J. (2018). Linear mixed-effects models and the analysis of nonindependent data: A unified framework to analyze categorical and continuous independent variables that vary within-subjects and/or within-items. *Psychological Methods*, *23*(3), 389–411. https://doi.org/10.1037/met0000159

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language*, *94*, 305–315. https://doi.org/10.1016/j.jml.2017.01.001

Singmann, H., & Kellen, D. (2019). An introduction to mixed models for experimental psychology. In D. Spieler & E. Schumacher (Eds.), *New methods in cognitive psychology* (pp. 4–31). Routledge. https://doi.org/10.4324/9780429318405-2
