---
title: Making research materials <b>F</b>indable, <b>A</b>ccessible, <b>I</b>nteroperable and <b>R</b>eusable
authors: ['<span style="font-size:140%;">Pablo Bernabeu, Gabriella Silva, Ngoc Giang My Hoang, Vincent DeLuca, Iva Ivanova, Jason Rothman, Jorge González Alonso</span>']
author: ~
date: '2024-03-08'
slug: making-research-materials-findable-accessible-interoperable-reusable-fair
categories:
  - research methods
tags:
  - materials
  - stimuli
  - open science
  - open data
  - reproducibility
  - research methods
  - psychology
  - linguistics
  - psycholinguistics
presentation_format: Poster
event: AcqVA Aurora Closing Event, UiT The Arctic University of Norway
event_url: 'https://site.uit.no/acqvalab/acqva-aurora-closing-event-8th-march-2024'
location: ~
address:
  street: ~
  city: ~
  region: ~
  postcode: ~
  country: ~
summary: ~
abstract: "To err is human, but when it comes to creating research materials, mistakes can be reduced by sharing more of our work and by using some helpful tools. These practices can make our research materials FAIRer—that is, more Findable, Accessible, Interoperable and Reusable (Wilkinson et al., 2016). In this presentation, I will describe how documentation practices and open-source software (namely, R and OpenSesame) were used in the preparation of materials for an artificial language study (for precedents, see Cross et al., 2021; González Alonso et al., 2020; Morgan-Short et al., 2012; Pereira Soares et al., 2022). Specifically, the use of documentation and open-source software facilitates the reproducibility, revision, modification and extension of the materials, both by the creators of the materials and by other researchers, at any point in time. I will describe how R was used during the preparation of linguistic and graphic stimuli for an experimental paradigm involving three artificial languages. The preparation of the stimuli drew on a modular framework formed of interoperable components (Figure 1). First, the minimal components of each language are contained in a base file in the ‘stimulus_preparation’ folder (Figure 2). Second, for the creation of the final stimuli, several controls were exerted to prevent spurious effects. For instance, gender and number were counterbalanced across experimental conditions, words were rotated across grammatical properties and sessions, and frequency of occurrence was controlled (Figure 3). The final stimuli can be recreated using the script 'compile_all_stimuli.R', and the resulting stimuli are saved to the 'session_materials' folder. Furthermore, I will describe how the current stimuli were presented using OpenSesame, which is a free-of-charge software program supporting experiments in the social sciences. I will describe the implementation of several complex sessions using OpenSesame, which involved the use of conditions for controlling the engagement of specific items (Figure 4). Last, I will describe a custom script that enables time-locking the EEG signal to the onset of specific stimuli, to measure event-related potentials (Figure 5)."
date_end: ~
all_day: no
publishDate: '2024-02-15T08:26:33+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
url_slides: ~
url_code: ~
url_pdf: ~
url_video: ~
projects: []
---


#### Figure 1

*Use of free, open-source, script-based software, such as R and OpenSesame, augments credibility and reliability of research.*

<div style = 'margin-top: 0; margin-bottom: 6%;'>

```{r, eval = FALSE, echo = TRUE}

└── stimulus_preparation
    ├── Norway site, base stimuli.csv
    ├── Spain site, base stimuli.csv
    ├── base_images.R
    ├── R_functions
    │   ├── Session2_Pretraining_vocabulary.R
    │   ├── Session2_Training_gender_agreement.R
    │   ├── Session2_Test_gender_agreement.R
    │   ├── Session2_Experiment_gender_agreement.R
    ...
    ├── compile_all_stimuli.R

```

</div>


#### Figure 2

*The minimal components of each language are contained in a base file.*

<div style = 'margin-top: 0; margin-bottom: 6%;'>

| verb_ID | verb_type   | verb       | verb_contrast_ID |
| ------- | ----------- | ---------- | ---------------- |
| 1       | copula_be   | is         |                  |
| 2       | copula_be   | are        |                  |
| 3       | copula_look | looks      |                  |
| 4       | copula_look | look       |                  |
| 5       | transitive  | remembered | A                |
| 6       | transitive  | forgot     | A                |
| 7       | transitive  | chose      | B                |
| 8       | transitive  | refused    | B                |

</div>


#### Figure 3

*Frequency of occurrence was controlled.*

<div style = 'margin-top: 0; margin-bottom: 6%;'>

```{r, eval = FALSE, echo = TRUE}

columns_to_check = c('noun1_gender', 'number', 'person', 
                     'verb', 'noun1', 'wrapup_noun')
for(i in seq_along(columns_to_check)) {
  column = columns_to_check[i]
  number_of_unique_frequencies = 
    combinations %>% 
    filter(complete.cases(get(column)), get(column) != '') %>% 
    group_by(get(column)) %>% tally() %>% select(n) %>% 
    n_distinct()
  if(number_of_unique_frequencies != 1) {
    warning(paste0('Some elements in the column `', column, 
                   '` appear more often than others.'))
    } }

```

</div>


#### Figure 4

*Complex sessions involving conditions for controlling the engagement of specific items.*

<div style = 'margin-top: 0; margin-bottom: 6%;'>
  <img style = 'margin-top: 0; margin-bottom: 6%;' src = 'images/Picture3.png' width = '95%'></img>
</div>


#### Figure 5

*Custom script for sending triggers in OpenSesame to support registration of ERPs.*

<div style = 'margin-top: 0; margin-bottom: 6%;'>
  <img style = 'margin-top: 0; margin-bottom: 6%;' src = 'images/Picture4.png' width = '95%'></img>
</div>


<div class='alert alert-warning' role='alert' style='padding-bottom: 0; margin-top: 7.5%; margin-bottom: 6%;'>

The poster is available to researchers and research students upon request while the study is in progress. Afterwards, it will be made public.

</div>


#### References

Cross, Z. R., Zou-Williams, L., Wilkinson, E. M., Schlesewsky, M., & Bornkessel-Schlesewsky, I. (2021). Mini Pinyin: A modified miniature language for studying language learning and incremental sentence processing. *Behavior Research Methods, 53(3)*, 1218–1239. https://doi.org/10.3758/s13428-020-01473-6

González Alonso, J., Alemán Bañón, J., DeLuca, V., Miller, D., Pereira Soares, S. M., Puig-Mayenco, E., Slaats, S., & Rothman, J. (2020). Event related potentials at initial exposure in third language acquisition: Implications from an artificial mini-grammar study. *Journal of Neurolinguistics, 56*, 100939. https://doi.org/10.1016/j.jneuroling.2020.100939

Morgan-Short, K., Finger, I., Grey, S., & Ullman, M. T. (2012). Second language processing shows increased native-like neural responses after months of no exposure. *PLOS ONE, 7*(3), e32974. https://doi.org/10.1371/journal.pone.0032974

Pereira Soares, S. M., Kupisch, T., & Rothman, J. (2022). Testing potential transfer effects in heritage and adult L2 bilinguals acquiring a mini grammar as an additional language: An ERP approach. *Brain Sciences, 12*(5), Article 5. https://doi.org/10.3390/brainsci12050669

Wilkinson, M. D., Dumontier, M., Aalbersberg, Ij. J., Appleton, G., Axton, M., Baak, A., Blomberg, N., Boiten, J.-W., da Silva Santos, L. B., Bourne, P. E., Bouwman, J., Brookes, A. J., Clark, T., Crosas, M., Dillo, I., Dumon, O., Edmunds, S., Evelo, C. T., Finkers, R., … Mons, B. (2016). The FAIR Guiding Principles for scientific data management and stewardship. *Scientific Data, 3*(1), Article 1. https://doi.org/10.1038/sdata.2016.18

