---
title: 'pilotr: Pilot the study before running it'
author: ''
date: '2026-09-01'
apa_captions: true
slug: pilotr-pilot-the-study-before-running-it
code_folding: hide
editable_r: true
categories:
  - R
  - statistics
  - research methods
tags:
  - s
  - R
  - power analysis
  - simulation
  - linear mixed-effects models
  - research methods
  - Python
  - psycholinguistics
subtitle: ''
summary: 'A simulated semantic-priming study shows how precision, sign errors, magnitude errors and singular fits can be examined alongside power before choosing a design.'
abstract: 'This tutorial uses a crossed semantic-priming design to show why simulation-based planning should examine more than power. A portable pilotr specification defines both the simulated data and the analysis model, and repeated simulation then reports precision, Type S (sign) error, Type M (magnitude) error and singular-fit warnings before any participants are recruited.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
package_docs:
  - name: R documentation
    url: https://pablobernabeu.github.io/pilotr/r/
  - name: Python documentation
    url: https://pablobernabeu.github.io/pilotr/python/
  - name: Browser app
    url: https://pablobernabeu.github.io/pilotr/app/
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, echo = FALSE,
  fig.align = 'center', fig.retina = 2, out.width = '100%',
  fig.width = 7.5, fig.height = 4.2, comment = '#>'
)
options(width = 90)
xaringanExtra::use_clipboard()
library(pilotr)
library(ggplot2)
workers <- 4L
n_sims <- 12L # An executable smoke test; planning runs need hundreds or more.

# Posts display figures at about 720 px, half their 1440 px width, so 1 pt
# reads as about 1.3 px on screen and axis text at 10 pt shows at about 13 px.
blog_theme <- theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 11),
    plot.title = element_text(size = 11.5, margin = margin(b = 8)),
    plot.title.position = 'plot',
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 14, 10, 10)
  )
percent_axis <- scales::label_percent(accuracy = 1)
# The orange used for fills and lines is too light for small text on white
# (3.9:1), so text labels use a darker shade of the same hue (5.3:1).
text_orange <- '#B34D00'
```

Power calculations can feel reassuringly final. Choose an effect, enter a sample size and read off a percentage. That simplicity disappears in experiments where participants respond to many items. Participants and items both vary, their responses are crossed, and the model usually needs random slopes as well as random intercepts. The design has become too structured for a single formula.

I built [pilotr](/software/pilotr/) (Bernabeu, 2026) to make the assumptions behind such a design visible before data collection. It stores a proposed study as a portable specification, simulates the observations that the specification implies and fits the planned analysis to them repeatedly. The R and Python packages read the same specification and generate the same data from the same seed. Below, I apply it to a semantic-priming study with 30 participants and 24 items to show what a design analysis has to weigh besides power.

## Development

The example moves from a transparent design specification to simulated observations and then to a set of planning diagnostics. Each analysis uses only 12 replicates, which keeps the code quick to run. The results therefore show the shape of each diagnostic and the quantity to read from it, but they cannot settle a sample size.

### A Study That Does Not Exist Yet

The imagined experiment is a lexical-decision task. Every participant sees every target word after both a related and an unrelated prime. Reaction times follow a shifted lognormal distribution, and the model includes item frequency, participant vocabulary and an interaction between vocabulary and priming. The main assumptions are set out in Table 1.

```{r design-assumptions}
knitr::kable(caption = 'The Assumptions Behind the Imagined Study.', data.frame(
  `Part of the design` = c('Participants', 'Items', 'Priming effect',
                           'Participant variation', 'Item variation', 'Response'),
  Assumption = c('30', '24', '0.05 on the log scale, about 20 ms',
                 'Random intercept and priming slope',
                 'Random intercept and priming slope',
                 'Shifted lognormal reaction time'),
  check.names = FALSE
), row.names = FALSE)
```

```{r define-spec}
spec <- list(
  spec_version = '0.3',
  name = 'semantic_priming_ldt',
  seed = 90210,
  units = list(subject = list(n = 30), item = list(n = 24)),
  factors = list(list(
    name = 'prime', levels = c('related', 'unrelated'),
    contrasts = list(prime = c(-0.5, 0.5)),
    vary_within = c('subject', 'item'))),
  predictors = list(
    list(name = 'z_freq', varies_by = 'item', mean = 0, sd = 1),
    list(name = 'z_vocab', varies_by = 'subject', mean = 0, sd = 1,
         reliability = 0.8)),
  fixed = list(
    intercept = 6,
    coefficients = list(prime = 0.05, z_freq = -0.03, z_vocab = -0.02,
                        `prime:z_vocab` = 0.02)),
  random = list(
    subject = list(intercept_sd = 0.12, slopes = list(prime = 0.04),
                   correlations = list(`intercept,prime` = 0.2)),
    item = list(intercept_sd = 0.08, slopes = list(prime = 0.02),
                correlations = list(`intercept,prime` = -0.1))),
  response = list(family = 'shifted_lognormal', name = 'RT', sigma = 0.30,
                  shift = 200, round = 4)
)
validate_spec(spec)
```

The specification determines the analysis formula as well as the data generator. That prevents a quiet mismatch between the design used for power and the model eventually fitted.

```{r show-formula, echo = TRUE}
model_formula(spec)
```

### Looking at the Imagined Data

A single simulated data set is useful before any power calculation (DeBruine & Barr, 2021). It shows whether the assumptions produce recognisable observations. In Figure 1, the simulated reaction times are right-skewed, as intended, and unrelated primes shift the distribution slightly to the right.

```{r simulate-data}
pilot <- simulate_design(spec)
```

```{r response-distribution, fig.height = 3.6, fig.cap = 'Simulated Reaction Times After Related and Unrelated Primes.', fig.alt = 'Overlapping density curves of simulated reaction times for related and unrelated primes, with the unrelated distribution shifted slightly to the right'}
# Label each curve directly instead of using a legend: related primes to
# the left of the rising edge near their peak, and unrelated primes to the
# right of the shoulder where that distribution stands furthest above the
# other. geom_density() uses the same kernel and bandwidth as density().
rt_range <- range(pilot$RT)
rt_density <- lapply(split(pilot$RT, pilot$prime), density,
                     from = rt_range[1], to = rt_range[2])
related_y <- 0.9 * max(rt_density$related$y)
related_edge <- which(rt_density$related$y >= related_y)[1]
unrelated_shoulder <- which.max(rt_density$unrelated$y - rt_density$related$y)
prime_labels <- data.frame(
  prime = c('related', 'unrelated'),
  RT = c(rt_density$related$x[related_edge] - 30,
         rt_density$unrelated$x[unrelated_shoulder] + 30),
  density = c(related_y, rt_density$unrelated$y[unrelated_shoulder]),
  label = c('Related primes', 'Unrelated primes'),
  hjust = c(1, 0)
)

ggplot(pilot, aes(RT, fill = prime)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_text(aes(RT, density, label = label, colour = prime, hjust = hjust),
            data = prime_labels, inherit.aes = FALSE, size = 10.5 / .pt) +
  scale_fill_manual(values = c(related = '#0072B2', unrelated = '#D55E00'),
                    guide = 'none') +
  scale_colour_manual(values = c(related = '#0072B2', unrelated = text_orange),
                      guide = 'none') +
  labs(x = 'Reaction time (ms)', y = 'Density') +
  blog_theme
```

The average priming effect is only part of the picture. The specification gives participants and items their own priming slopes, and trial-level noise adds to that variation, so some show a much larger difference than others. Figure 2 gives each unit its own row, and in about a third of each set the difference runs the other way.

```{r unit-effects, fig.height = 4.4, fig.cap = 'Simulated Priming Effects for Each Item and Each Participant.', fig.alt = 'Dot plots of simulated priming effects for each item and participant, sorted within two panels, with most effects above zero and about a third below it'}
unit_effects <- function(unit) {
  means <- tapply(pilot$RT, list(pilot[[unit]], pilot$prime), mean)
  data.frame(unit = unit, id = rownames(means),
             effect_ms = means[, 'unrelated'] - means[, 'related'])
}
effects <- rbind(unit_effects('item'), unit_effects('subject'))
effects$id <- factor(
  paste(effects$unit, effects$id),
  levels = paste(effects$unit, effects$id)[order(effects$unit, effects$effect_ms)]
)

ggplot(effects, aes(effect_ms, id, colour = unit)) +
  geom_vline(xintercept = 0, colour = 'grey60') +
  geom_point(size = 2.2) +
  facet_wrap(
    ~ unit, scales = 'free_y',
    labeller = as_labeller(c(item = 'Items', subject = 'Participants'))
  ) +
  scale_colour_manual(values = c(item = '#D55E00', subject = '#0072B2'),
                      guide = 'none') +
  labs(x = 'Priming effect (unrelated minus related trials, ms)', y = NULL) +
  blog_theme +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank(),
        panel.spacing.x = unit(1.5, 'lines'))
```

### Detection Is Not Estimation

The first design analysis simulates the study repeatedly, fits the mixed model to each simulated data set, and records conventional power and the number of singular fits. It also reports Type S error, the chance that a significant estimate has the wrong sign, and Type M error, the factor by which significant estimates exaggerate the true effect (Gelman & Carlin, 2014).

```{r planned-power, echo = TRUE}
power_at_plan <- power_mixed(spec, n_sims = n_sims, workers = workers)
power_at_plan
```

With 12 replicates, a single simulated success moves a power figure by more than 8 percentage points. A planning run would use the same code with several hundred replicates, enough to bring the Monte Carlo standard error in the `mcse` column down to an acceptable level. The power curve in Figure 3 comes from repeating the analysis at six sample sizes, from 16 to 96 participants.

```{r power-curve-data}
power_curve <- power_curve_mixed(
  spec,
  subject_ns = c(16, 24, 32, 48, 64, 96),
  n_sims = n_sims,
  workers = workers
)
priming_curve <- power_curve[power_curve$effect == 'prime', ]
```

```{r power-curve, fig.height = 4.2, fig.cap = sprintf('Detection Probability for the Priming Effect Across Participant Counts. Each design uses %s replicates. All simulated designs contain 24 items.', n_sims), fig.alt = 'Illustrative power estimates for the priming effect, rising from about 17% with 16 participants to 100% with 96 and crossing the 80% reference line between 32 and 48 participants'}
ggplot(priming_curve, aes(n_subject, power)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = 'grey55') +
  geom_ribbon(aes(ymin = power_lo, ymax = power_hi),
              alpha = 0.15, fill = '#0072B2') +
  geom_line(colour = '#0072B2', linewidth = 0.8) +
  geom_point(colour = '#0072B2', size = 2.4) +
  scale_x_continuous(breaks = priming_curve$n_subject) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = percent_axis) +
  labs(x = 'Number of participants', y = 'Power for the priming effect') +
  blog_theme
```

Power asks whether an effect would clear a threshold. I also want to know whether its interval would exclude effects too small to matter. With a region of practical equivalence (ROPE; Kruschke, 2018) of plus or minus 0.02 on the log scale, about 8 ms here, precision becomes a separate design criterion. Across samples of 16 to 200 participants, the left panel of Figure 4 shows how often the 95% confidence interval excludes the ROPE, and the right panel shows the interval's mean width. The dashed line in the right panel marks a width of 0.06, below which an interval centred on the true effect of 0.05 would exclude the ROPE.

```{r precision-data}
precision <- precision_curve(
  spec,
  focal = c(prime = 0.05),
  subject_ns = c(16, 32, 64, 96, 128, 160, 200),
  rope = 0.02,
  n_sims = n_sims,
  workers = workers
)
```

```{r precision-curve, fig.height = 3.8, fig.cap = sprintf('Precision, as the Probability That an Interval Excludes a Negligible Effect and as the Mean Width of That Interval, Across Participant Counts. Each design uses %s replicates. All simulated designs contain 24 items. In the panel titles, CI stands for confidence interval, ROPE for region of practical equivalence and RT for reaction time.', n_sims), fig.alt = 'Two panels of illustrative estimates showing the probability that an interval excludes negligible effects and its mean width as the number of participants increases'}
# Rotated beside panels this short, the long y-axis descriptions would run
# the full panel height, so each becomes its panel's title instead. The
# shared x-axis title is drawn once.
p_probability <- ggplot(precision, aes(n_subject, p_meaningful)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = 'grey55') +
  geom_ribbon(aes(ymin = p_meaningful_lo, ymax = p_meaningful_hi),
              alpha = 0.15, fill = '#0072B2') +
  geom_line(colour = '#0072B2', linewidth = 0.8) +
  geom_point(colour = '#0072B2', size = 2.2) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = percent_axis) +
  labs(x = 'Number of participants', y = NULL,
       title = 'Probability the 95% CI excludes the ROPE') +
  blog_theme +
  theme(plot.margin = margin(0, 30, 0, 10))

p_width <- ggplot(precision, aes(n_subject, mean_ci_width)) +
  geom_hline(yintercept = 2 * (0.05 - 0.02), linetype = 2, colour = 'grey55') +
  geom_line(colour = '#0072B2', linewidth = 0.8) +
  geom_point(colour = '#0072B2', size = 2.2) +
  scale_y_continuous(breaks = seq(0.03, 0.12, 0.03)) +
  labs(x = 'Number of participants', y = NULL,
       title = 'Mean 95% CI width (log-RT units)') +
  blog_theme +
  theme(plot.margin = margin(0, 14, 0, 10))

patchwork::wrap_plots(p_probability, p_width) +
  patchwork::plot_layout(axis_titles = 'collect')
```

### The Significant Estimates Are the Misleading Ones

A low-powered study misses real effects, and it also distorts the ones it finds. When power is low, only unusually large estimates pass the significance threshold, so significance selects for exaggeration. To see that selection directly, I varied the true effect with 30 participants and set the result, in Figure 5, beside the sample-size curve from above, in which the effect stays at 0.05.

```{r type-m-data}
effect_grid <- design_conditions(
  prime = c(0.03, 0.05, 0.08),
  .null = TRUE,
  .base = list(z_freq = -0.03, z_vocab = -0.02, `prime:z_vocab` = 0.02)
)
effect_sweep <- sweep_spec(
  spec, 'fixed$coefficients', effect_grid, power_mixed,
  n_sims = n_sims, workers = workers, .name = 'condition'
)
effect_sweep <- effect_sweep[effect_sweep$effect == 'prime', ]
type_m <- rbind(
  data.frame(series = 'Effect size (30 participants)', power = effect_sweep$power,
             type_m = effect_sweep$type_m),
  data.frame(series = 'Sample size (effect = 0.05)', power = priming_curve$power,
             type_m = priming_curve$type_m)
)
type_m <- type_m[is.finite(type_m$type_m), ]
```

```{r type-m-plot, fig.height = 4.4, fig.cap = sprintf('Magnitude Exaggeration Among the Statistically Significant Estimates, Against Power. Each design uses %s replicates.', n_sims), fig.alt = 'Type M error falls from about 1.4 and 1.2 at low power to about 1 at high power in both an effect-size sweep and a sample-size curve, crossing slightly below the no-exaggeration line'}
# Each series is labelled beside its lowest-power point instead of in a
# legend: the steep effect-size series just above and to the right of it,
# and the flatter sample-size series below its first segment.
series_labels <- do.call(rbind, lapply(
  split(type_m, type_m$series), function(d) d[which.min(d$power), ]
))
# Rows follow the alphabetical series order: effect size, then sample size.
series_labels$power <- series_labels$power + c(0.015, -0.005)
series_labels$type_m <- series_labels$type_m + c(0.01, -0.07)
series_labels$vjust <- c(0, 1)
series_labels$text_colour <- c('#0072B2', text_orange)

ggplot(type_m, aes(power, type_m, colour = series)) +
  geom_hline(yintercept = 1, linetype = 2, colour = 'grey55') +
  annotate('text', x = max(type_m$power), y = 1, vjust = -0.6, hjust = 1,
           label = 'No average exaggeration', colour = 'grey35',
           size = 9.5 / .pt) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  geom_text(aes(label = series, vjust = vjust), data = series_labels,
            colour = series_labels$text_colour, hjust = 0,
            size = 10.5 / .pt) +
  scale_colour_manual(values = c('#0072B2', '#D55E00'), guide = 'none') +
  scale_x_continuous(breaks = seq(0, 1, 0.2), labels = percent_axis) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(x = 'Power', y = 'Type M error') +
  blog_theme
```

Type M error falls as power rises in both series, from about 1.4 (effect size) and 1.2 (sample size) at the lowest power to roughly 1 at the highest. With 12 replicates, it also wanders below 1, which a planning run with several hundred replicates would not.

Together, these diagnostics cover what a planning run has to examine: detection, precision, magnitude error and singular fits. The frequent singular fits in this example call for a second look at the random-effects structure before anyone is recruited. Keeping that structure maximal protects the Type I error rate (Barr et al., 2013), and trimming it when the data cannot support it protects power (Matuschek et al., 2017). A planning run lets me weigh that trade-off while the design can still change.

## Discussion

The plots should be read as a sensitivity analysis over assumptions that the real data may not bear out. Their practical purpose is to reveal which design feature limits the study: too few participants, too much unit-level variation, an interval that remains too wide or a model that frequently becomes singular. A defensible decision records those trade-offs and tests plausible alternatives, because the first design to cross an 80% power line may still estimate the effect imprecisely or exaggerate it.

The R package is available from the Comprehensive R Archive Network (CRAN) and the Python package from the Python Package Index (PyPI). The [R documentation](https://pablobernabeu.github.io/pilotr/r/) and [Python documentation](https://pablobernabeu.github.io/pilotr/python/) describe the complete specification format, response families and mixed-model options. For planning without code, the [browser app](https://pablobernabeu.github.io/pilotr/app/) builds and simulates a design and estimates power for a simple two-group comparison, while the mixed-effects and precision analyses shown here run in the installed R package.

## References

Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language*, *68*(3), 255–278. https://doi.org/10.1016/j.jml.2012.11.001

Bernabeu, P. (2026). *pilotr: Simulate experimental and behavioural data from a portable design specification* (Version 0.3.0) [Computer software]. CRAN. https://doi.org/10.32614/CRAN.package.pilotr

DeBruine, L. M., & Barr, D. J. (2021). Understanding mixed-effects models through data simulation. *Advances in Methods and Practices in Psychological Science*, *4*(1), Article 2515245920965119. https://doi.org/10.1177/2515245920965119

Gelman, A., & Carlin, J. (2014). Beyond power calculations: Assessing Type S (sign) and Type M (magnitude) errors. *Perspectives on Psychological Science*, *9*(6), 641–651. https://doi.org/10.1177/1745691614551642

Kruschke, J. K. (2018). Rejecting or accepting parameter values in Bayesian estimation. *Advances in Methods and Practices in Psychological Science*, *1*(2), 270–280. https://doi.org/10.1177/2515245918771304

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language*, *94*, 305–315. https://doi.org/10.1016/j.jml.2017.01.001
