---
title: 'pilotr: pilot the study before running it'
author: ''
date: '2026-09-01'
slug: pilotr-pilot-the-study-before-running-it
categories:
  - statistics
  - research methods
tags:
  - power analysis
  - simulation
  - linear mixed-effects models
  - research methods
  - Python
  - psycholinguistics
subtitle: ''
summary: 'A study with participants and items crossed has a power that no formula gives, and the only way to learn it is to simulate the design. pilotr describes such a design once, as a portable specification, and simulates the data it implies, in R and in Python, from the same bytes. This post pilots a semantic-priming experiment before it exists: the reaction times it would produce, the power of its planned analysis, the exaggeration of the effects that would reach significance, the sample size that would make the effect distinguishable from a negligible one, and the artefacts that make the plan reproducible.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
knitr::opts_chunk$set(python.reticulate = FALSE, engine.path = list(
  python = '/tmp/claude-0/-home-user/22b7e646-ecfa-5c99-a752-a7e5438334ea/scratchpad/pyenv/bin/python'
))
options(width = 95)
xaringanExtra::use_clipboard()
out_dir <- file.path(tempdir(), 'pilotr-post')
dir.create(out_dir, showWarnings = FALSE)
Sys.setenv(PILOTR_POST_OUT = out_dir)
```

Most experiments in psycholinguistics cross participants with items, and the analysis that respects that structure is a mixed model with random intercepts and slopes for both (Barr et al., 2013). No formula gives the power of such an analysis. The power depends on the number of participants and of items, on the variance of each source of variation, on the correlations between intercepts and slopes and on the response distribution, and the only way to learn it is to write down the design and simulate it many times (DeBruine & Barr, 2021). Simulating before the study, while the design can still be changed, is what makes the chosen sample size a decision. It is also a chance to see the data that the design would produce, and to catch a mistaken assumption while it is cheap.

In pilotr, the design is the object that is written, validated and reused. A design is a specification, a small JSON document with the units, the factors, the predictors, the fixed effects, the random effects and the response family, which is validated before anything is simulated. The same specification produces a data set, the formula of the analysis it implies, a power analysis with Type S and Type M errors (Gelman & Carlin, 2014), a precision analysis against a region of practical equivalence, and a set of scripts that reproduce the whole exercise. The R and the Python packages read the same specification and, given the same seed, generate the same data, down to the bytes.

```{r}
library(pilotr)
library(ggplot2)
packageVersion('pilotr')
workers <- 4L
n_sims <- 50L
```

## The design, written once

The experiment is a semantic-priming lexical-decision task: 30 participants, 24 target words, each seen with a related and an unrelated prime. Reaction times follow a shifted lognormal distribution, 200 ms plus a lognormal component, so the effects are on the log scale, where an intercept of 6 puts the typical response near 600 ms and a priming effect of 0.05 is a slowing of about 20 ms for unrelated primes. Two continuous predictors enter, the frequency of the target word at the item level and the vocabulary of the participant at the participant level, the latter measured with a reliability of 0.8, which is what a vocabulary test delivers. A cross-level interaction lets priming grow with vocabulary. Both participants and items have random intercepts and random priming slopes, with their correlations.

```{r}
spec <- list(
  spec_version = '0.3',
  name = 'semantic_priming_ldt',
  seed = 90210,
  units = list(subject = list(n = 30), item = list(n = 24)),
  factors = list(list(
    name = 'prime', levels = c('related', 'unrelated'),
    contrasts = list(prime = c(-0.5, 0.5)),
    vary_within = c('subject', 'item'))),
  predictors = list(
    list(name = 'z_freq',  varies_by = 'item',    mean = 0, sd = 1),
    list(name = 'z_vocab', varies_by = 'subject', mean = 0, sd = 1, reliability = 0.8)),
  fixed = list(
    intercept = 6.0,
    coefficients = list(prime = 0.05, z_freq = -0.03, z_vocab = -0.02,
                        `prime:z_vocab` = 0.02)),
  random = list(
    subject = list(intercept_sd = 0.12, slopes = list(prime = 0.04),
                   correlations = list(`intercept,prime` = 0.2)),
    item    = list(intercept_sd = 0.08, slopes = list(prime = 0.02),
                   correlations = list(`intercept,prime` = -0.1))),
  response = list(family = 'shifted_lognormal', name = 'RT', sigma = 0.30,
                  shift = 200, round = 4))
validate_spec(spec)
```

Validation is there because a mistyped key would otherwise be ignored, and the design would be simulated without the effect it was meant to test.

```{r}
bad <- spec
names(bad$fixed$coefficients)[1] <- 'prme'
tryCatch(validate_spec(bad), error = function(e) conditionMessage(e))
```

## What the data would look like

A single simulated data set is the first thing to inspect, and often the most instructive, since it shows whether the assumed variances produce reaction times that look like reaction times.

```{r}
pilot <- simulate_design(spec)
dim(pilot)
head(pilot)
tapply(pilot$RT, pilot$prime, mean)
```

```{r, fig.width = 7, fig.height = 3.6, fig.alt = 'Overlapping density curves of simulated reaction times for related and unrelated primes, right-skewed, with the unrelated curve shifted to the right'}
ggplot(pilot, aes(RT, fill = prime)) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_manual(values = c(related = '#0072B2', unrelated = '#D55E00')) +
  labs(x = 'RT (ms)', y = 'Density', fill = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = 'top')
```

Because the specification gave both participants and items random priming slopes, each participant and each item has its own priming effect, which is the variation that the analysis has to account for and that a formula ignores.

```{r, fig.width = 7.5, fig.height = 3.8, fig.alt = 'Dot plots of the priming effect in milliseconds for each participant and for each item, each panel sorted by effect size, spread on both sides of zero'}
unit_effects <- function(unit) {
  m <- tapply(pilot$RT, list(pilot[[unit]], pilot$prime), mean)
  data.frame(unit = unit, id = rownames(m), effect_ms = m[, 'unrelated'] - m[, 'related'])
}
effs <- rbind(unit_effects('subject'), unit_effects('item'))
# Subject and item identifiers both run from 1, so they are prefixed before
# being ordered by effect size, or the levels of one unit would order the other.
effs$id <- paste(effs$unit, effs$id)
effs$id <- factor(effs$id, levels = effs$id[order(effs$unit, effs$effect_ms)])
ggplot(effs, aes(effect_ms, id, colour = unit)) +
  geom_vline(xintercept = 0, colour = 'grey60') +
  geom_point(size = 2) +
  facet_wrap(~ unit, scales = 'free_y') +
  scale_colour_manual(values = c(subject = '#0072B2', item = '#D55E00'), guide = 'none') +
  labs(x = 'Priming effect (ms), unrelated minus related', y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
```

## The anatomy of the design

The specification implies a decomposition of the variance of the linear predictor and an analysis model, and both can be read off it before any data exist.

```{r}
round(unlist(response_variance(spec)), 4)
model_formula(spec)
```

Whether a simulation recovers the parameters that went into it is the check that DeBruine and Barr (2021) recommend before trusting a power analysis. One large simulated study, fitted with the implied model, returns the fixed effects below alongside their standard errors and the distance of each estimate from the value that was specified.

```{r}
big <- spec
big$units$subject$n <- 80
big$units$item$n <- 60
fit <- suppressMessages(lmerTest::lmer(
  model_formula(spec), data = model_data(spec, simulate_design(big)),
  control = lme4::lmerControl(calc.derivs = FALSE)))
truth <- c(`(Intercept)` = 6, prime = 0.05, z_freq = -0.03, z_vocab = -0.02,
           prime_z_vocab = 0.02)
est <- lme4::fixef(fit)[names(truth)]
se  <- sqrt(diag(as.matrix(vcov(fit))))[names(truth)]
data.frame(specified = truth, recovered = round(est, 4), se = round(se, 4),
           # Distance from the specified value, in standard errors.
           z = round((est - truth) / se, 2))
```

The two vocabulary terms come out smaller in magnitude than specified, the direction that measuring vocabulary with a reliability of 0.8 predicts. A single replicate cannot separate that attenuation from sampling noise, though, since the largest distance in the last column is `r sprintf('%.2f', max(abs((est - truth) / se)))` standard errors and every estimate is therefore consistent with the value it was given.

The reverse direction is how a real pilot seeds a power analysis: a specification is read off a fitted model and scaled to the planned sample.

```{r}
from_fit <- spec_from_model(fit, name = 'from_pilot_fit', n_subject = 60, n_item = 40,
                            family = list(family = 'shifted_lognormal', shift = 200),
                            round = 4)
c(subject_intercept_sd = from_fit$random$subject$intercept_sd,
  subject_prime_slope = from_fit$random$subject$slopes$prime,
  sigma = from_fit$response$sigma)
```

## Power, and what a significant result would look like

Simulation-based power at the planned sample refits the implied model on each replicate and counts how often each effect reaches significance. The report includes the Monte Carlo standard error, a Wilson interval, the number of fits that were singular, which are counted and not dropped, and for each effect the Type S error, the share of significant estimates with the wrong sign, and the Type M error, the average factor by which significant estimates exaggerate the true effect.

```{r}
pw <- power_mixed(spec, n_sims = n_sims, workers = workers)
pw
```

With 30 participants and 24 items, the maximal model is richer than the design supports, and many replicates end in a boundary fit, the situation in which Matuschek et al. (2017) show that the cost in power outweighs the protection against Type I error. The priming effect has reasonable power. The interaction with vocabulary, which is what a researcher would most want to claim, does not, and the estimates of it that would reach significance are exaggerated by a factor of two or more. A sweep over effect sizes, with a null condition, shows the false-positive rate of the planned analysis and how the exaggeration recedes as power grows. The base coefficients are carried along explicitly, since the sweep replaces the block of coefficients as a whole.

```{r}
grid <- design_conditions(prime = c(0.03, 0.05, 0.08), .null = TRUE,
                          .base = list(z_freq = -0.03, z_vocab = -0.02,
                                       `prime:z_vocab` = 0.02))
sweep <- sweep_spec(spec, 'fixed$coefficients', grid, power_mixed,
                    n_sims = n_sims, workers = workers, .name = 'condition')
sweep[sweep$effect == 'prime',
      c('condition', 'true', 'power', 'power_lo', 'power_hi', 'type_s', 'type_m',
        'n_singular')]
```

## How many participants

A binomial regression of power on the square root of the sample size, weighted by the replicates behind each point, solves the curve for the sample that reaches 80% power, with an interval from the delta method. The solver returns no estimate outside the simulated range.

```{r}
curve <- power_curve_mixed(spec, subject_ns = c(16, 24, 32, 48, 64, 96),
                           n_sims = n_sims, workers = workers)
cv_prime <- curve[curve$effect == 'prime', ]
cv_prime[, c('n_subject', 'power', 'power_lo', 'power_hi', 'type_m', 'n_singular')]
solved <- target_n(curve, target = 0.8, effect = 'prime')
c(n = solved$n, n_lo = solved$n_lo, n_hi = solved$n_hi)
```

```{r, fig.width = 7, fig.height = 4, fig.alt = 'Power curve for the priming effect from 16 to 96 participants with a Monte Carlo band, the 80% line and the solved sample size marked'}
ggplot(cv_prime, aes(n_subject, power)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = 'grey60') +
  annotate('rect', xmin = solved$lo, xmax = solved$hi, ymin = -Inf, ymax = Inf,
           fill = 'grey60', alpha = 0.15) +
  geom_vline(xintercept = solved$value, linetype = 2, colour = 'grey60') +
  geom_ribbon(aes(ymin = power_lo, ymax = power_hi), alpha = 0.15, fill = '#0072B2') +
  geom_line(colour = '#0072B2', linewidth = 0.8) +
  geom_point(colour = '#0072B2', size = 2.6) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Participants (24 items)', y = 'Power for the priming effect') +
  theme_minimal(base_size = 13)
```

Power concerns detection. Precision concerns whether the study could tell a priming effect from a negligible one, which is what a region of practical equivalence (ROPE) formalises (Kruschke, 2018). With a ROPE of ±0.02 on the log scale, about 8 ms, the precision analysis asks how often the 95% interval would fall entirely outside the region, and solves the curve for the sample that reaches 80%.

```{r}
prc <- precision_curve(spec, focal = c(prime = 0.05),
                       subject_ns = c(16, 32, 64, 96, 128, 160, 200),
                       rope = 0.02, n_sims = n_sims, workers = workers)
prc[, c('n_subject', 'mean_ci_width', 'p_meaningful', 'p_meaningful_lo',
        'p_meaningful_hi', 'n_singular')]
solved_prec <- solve_curve(prc, target = 0.8)
round(c(n = solved_prec$value, lo = solved_prec$lo, hi = solved_prec$hi))
```

```{r, fig.width = 8, fig.height = 3.6, fig.alt = 'Two panels: the probability that the 95% interval lies outside the region of practical equivalence, and the mean interval width, both against the number of participants'}
p_left <- ggplot(prc, aes(n_subject, p_meaningful)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = 'grey60') +
  annotate('rect', xmin = solved_prec$lo, xmax = solved_prec$hi, ymin = -Inf, ymax = Inf,
           fill = 'grey60', alpha = 0.15) +
  geom_ribbon(aes(ymin = p_meaningful_lo, ymax = p_meaningful_hi),
              alpha = 0.15, fill = '#0072B2') +
  geom_line(colour = '#0072B2', linewidth = 0.8) + geom_point(colour = '#0072B2', size = 2.2) +
  labs(x = 'Participants (24 items)', y = 'P(95% CI outside the ROPE)') +
  theme_minimal(base_size = 12)
p_right <- ggplot(prc, aes(n_subject, mean_ci_width)) +
  geom_hline(yintercept = 2 * (0.05 - 0.02), linetype = 2, colour = 'grey60') +
  geom_line(colour = '#0072B2', linewidth = 0.8) + geom_point(colour = '#0072B2', size = 2.2) +
  labs(x = 'Participants (24 items)', y = 'Mean 95% CI width (log scale)') +
  theme_minimal(base_size = 12)
patchwork::wrap_plots(p_left, p_right)
```

The sample that detects the effect is well below the sample that estimates it precisely, which is the usual finding and the reason to plan for precision when the size of an effect matters more than its existence.

```{r, fig.width = 7, fig.height = 4, fig.alt = 'Type M error against power from the effect-size sweep and the sample-size curve, falling towards one as power rises'}
sw_prime <- sweep[sweep$effect == 'prime', ]
tm <- rbind(data.frame(source = 'effect-size sweep, N = 30', power = sw_prime$power,
                       type_m = sw_prime$type_m),
            data.frame(source = 'sample-size curve, effect = 0.05', power = cv_prime$power,
                       type_m = cv_prime$type_m))
tm <- tm[is.finite(tm$type_m), ]
ggplot(tm, aes(power, type_m, colour = source)) +
  geom_hline(yintercept = 1, linetype = 2, colour = 'grey60') +
  geom_line() + geom_point(size = 2.4) +
  scale_colour_manual(values = c('#0072B2', '#D55E00')) +
  labs(x = 'Power', y = 'Type M error (exaggeration of significant estimates)',
       colour = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = 'top')
```

## The artefacts

The artefacts below let someone else rerun the plan. The specification is written as JSON, an R script that embeds it at full precision is generated, the confirmatory brms model that the specification implies is written out with its formula, family and priors, and a Bayesian design analysis with a Savage-Dickey Bayes factor and a ROPE decision can be emitted as a script for a cluster. The replicate seeds come from a generator shared with the Python package.

```{r}
writeLines(spec_json(spec), file.path(out_dir, 'semantic_priming_ldt.json'))
writeLines(generate_r_script(spec), file.path(out_dir, 'reproduce_design.R'))
invisible(brms_bridge(spec))
replicate_seeds(spec$seed, 5)
```

## The same bytes from Python

The two packages implement the same generator over the same specification, down to the order in which random deviates are drawn. Below, the R session writes the simulated pilot data with every value at 17 significant digits and hashes it. The Python session then reads the specification, simulates and hashes the same format. The function names differ slightly, `simulate_design()` in R against `simulate()` in Python, since each follows its own language's conventions.

```{r}
cell <- function(v) if (is.character(v)) v else sprintf('%.17g', as.numeric(v))
dump17 <- function(d, path) {
  con <- file(path, open = 'wb'); on.exit(close(con))
  cols <- lapply(d, function(col) vapply(col, cell, character(1), USE.NAMES = FALSE))
  lines <- c(paste(names(d), collapse = ','),
             vapply(seq_len(nrow(d)), function(i)
               paste(vapply(cols, `[`, character(1), i), collapse = ','), character(1)))
  writeLines(lines, con, sep = '\n')
}
dump17(pilot, file.path(out_dir, 'r_dump.txt'))
unname(tools::md5sum(file.path(out_dir, 'r_dump.txt')))
```

```{python}
import hashlib, json, os
import pilotr

out = os.environ['PILOTR_POST_OUT']
with open(os.path.join(out, 'semantic_priming_ldt.json')) as fh:
    spec = json.load(fh)
pilotr.validate_spec(spec)
data = pilotr.simulate(spec)

def cell(v):
    return v if isinstance(v, str) else format(float(v), '.17g')

lines = [','.join(data.columns)]
lines += [','.join(cell(row[c]) for c in data.columns) for row in data.rows]
text = '\n'.join(lines) + '\n'
print('pilotr', pilotr.__version__)
print(hashlib.md5(text.encode('utf-8')).hexdigest())
print('replicate seeds:', pilotr.replicate_seeds(spec['seed'], 5))
```

`r n_sims` replicates per point are enough to show the shape of the curves in a post and far too few for a plan, where two hundred or more are the norm and the Monte Carlo standard error is the quantity to watch. The specification also encodes assumptions, in particular the random-effect variances and correlations, that a real pilot or a published data set should supply, and the reverse direction from a fitted model exists for that reason.

### References

Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language*, *68*(3), 255–278. https://doi.org/10.1016/j.jml.2012.11.001

DeBruine, L. M., & Barr, D. J. (2021). Understanding mixed-effects models through data simulation. *Advances in Methods and Practices in Psychological Science*, *4*(1). https://doi.org/10.1177/2515245920965119

Gelman, A., & Carlin, J. (2014). Beyond power calculations: Assessing Type S (sign) and Type M (magnitude) errors. *Perspectives on Psychological Science*, *9*(6), 641–651. https://doi.org/10.1177/1745691614551642

Kruschke, J. K. (2018). Rejecting or accepting parameter values in Bayesian estimation. *Advances in Methods and Practices in Psychological Science*, *1*(2), 270–280. https://doi.org/10.1177/2515245918771304

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language*, *94*, 305–315. https://doi.org/10.1016/j.jml.2017.01.001
