---
title: Speculation across various scientific topics
author: ''
date: '2026-09-02'
slug: speculation-across-various-scientific-topics
categories:
  - science
  - bibliometrics
tags:
  - linguistics
  - language
  - bibliometrics
  - Scopus
  - research methods
subtitle: ''
summary: 'Scopus searches comparing, year by year since 1980, the share of publications that mention speculation within nine literatures: language evolution, language comprehension, language disorders, linguistic relativity, language teaching, the bilingual advantage, sensorimotor simulation, artificial general intelligence and the hadron collider. The searches are run through the scopusflow package by a workflow that anyone with a key can rerun.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 95)
xaringanExtra::use_clipboard()
```

<div class='alert alert-info' role='alert' style='padding-bottom: 0; margin-bottom: 6%;'>

Given all of the data that E&L [Evans & Levinson, 2009] cite, how could anyone maintain the notion of a universal grammar with linguistic content? Traditionally, there have been three basic strategies. First, just as we may force English grammar into the Procrustean bed of Latin grammar – that is how I was taught the structure of English in grade school – the grammars of the world's so-called exotic languages may be forced into an abstract scheme based mainly on European languages.

— Tomasello (2009)

</div>

Some research topics invite speculation more than others. When Pieter Vogt and I reviewed the field of language evolution a decade ago, the impression was of a literature in which conjecture outran evidence by some distance, for the understandable reason that the events under study left no direct trace (Bernabeu & Vogt, 2015). The passage from Tomasello quoted above makes a similar complaint about universal grammar. An impression of this kind can be checked, at least crudely, by counting. Authors who speculate in print often say so, in phrases such as 'we speculate that' or 'it is tempting to speculate', and those phrases are indexed when they occur in a title, an abstract or a keyword list. The share of records in a literature that contain a form of the word 'speculate' is therefore a rough measure of how openly conjectural that literature is, or at least of how often its authors label their conjectures as such.

This post compares that share, year by year from 1980 to the present, across nine reference literatures. Six of them concern language: language evolution, language comprehension, language disorders, linguistic relativity, language teaching and the bilingual advantage. A seventh, sensorimotor simulation, is the topic of my own doctoral work on conceptual processing. The last two, artificial general intelligence and the hadron collider, sit outside the language sciences and serve as points of comparison, one being a field where speculation might be expected and the other a field of large instruments and large collaborations. Each reference literature is defined by a query on the title, abstract and keywords, and the comparison counts, within each year, the records that also match `speculat*` in the same fields.

## The searches

An earlier draft of this post ran the searches through a set of functions that I had written around the rscopus package. Those functions have since become the [scopusflow package](/2026/scopusflow-a-literature-search-you-can-rerun/), and the searches now live in a script in the [website's repository](https://github.com/pablobernabeu/website-files/blob/master/scripts/update_post_searches.R) that a GitHub Actions workflow runs on request, with the Scopus key held as a repository secret. The workflow commits the counts to the post's directory, so the figures below can be rebuilt from the committed data without a key, and the searches can be refreshed by anyone with a key by dispatching the workflow. The nine queries are given below.

```{r, eval = FALSE}
library(scopusflow)

years <- 1980:as.integer(format(Sys.Date(), '%Y'))

topics <- c(
  'language evolution'              = '"language evolution" OR "evolution of language"',
  'language comprehension'          = '"language comprehension"',
  'language disorders'              = '"language disorders" OR "language disorder"',
  'linguistic relativity'           = '"linguistic relativity"',
  'language teaching'               = '"language teaching"',
  'bilingual advantage'             = '"bilingual advantage" OR "bilingual advantages"',
  'sensorimotor simulation'         = '"sensorimotor simulation"',
  'artificial general intelligence' = '"artificial general intelligence"',
  'hadron collider'                 = '"hadron collider"'
)

comparisons <- lapply(names(topics), function(label) {
  cmp <- scopus_compare_topics(
    reference_query  = topics[[label]],
    comparison_terms = 'speculat*',
    years            = years,
    field            = 'TITLE-ABS-KEY'
  )
  cmp$topic <- label
  cmp
})
```

The function counts, for every year, the records matching the reference query and the records matching both the reference query and the comparison term, so the nine comparisons cost `r 9 * 2 * length(unique(readRDS('searches/speculation_comparisons.rds')[[1]]$year))` count requests. Each row of the result holds a year, the count for the reference literature, the count for the comparison and the comparison as a percentage of the reference.

```{r}
library(dplyr)
library(ggplot2)

comparisons <- readRDS('searches/speculation_comparisons.rds')
retrieved <- attr(comparisons, 'retrieved')
retrieved

speculation <- bind_rows(comparisons) |>
  filter(query_type == 'comparison') |>
  transmute(topic, year, reference_n, n,
            percentage = comparison_percentage)
head(speculation)
```

The searches were run on `r format(as.Date(substr(retrieved, 1, 10)), '%d %B %Y')`, so the final year is incomplete and its share rests on fewer records than the years before it.

## Sizes of the literatures

The nine literatures differ in size by orders of magnitude, and the size matters for reading the percentages. A share computed on a few dozen records moves by several points when one or two abstracts mention speculation, whereas a share computed on thousands of records barely moves at all.

```{r}
totals <- speculation |>
  group_by(topic) |>
  summarise(records = sum(reference_n), with_speculation = sum(n),
            share = 100 * with_speculation / records,
            first_year_with_records = min(year[reference_n > 0])) |>
  arrange(desc(share))
totals
```

Across the whole period, the share of records that mention speculation is highest in the literature on `r totals$topic[1]` (`r sprintf('%.1f', totals$share[1])`%) and lowest in `r totals$topic[nrow(totals)]` (`r sprintf('%.1f', totals$share[nrow(totals)])`%), with the language literatures spread between those two.

```{r, fig.width = 8, fig.height = 4, fig.alt = 'Line chart on a logarithmic axis of the number of Scopus records per year in each of the nine reference literatures from 1980 onwards'}
# Nine literatures against an eight-colour palette, so the Okabe-Ito set is
# extended with one grey. Nine lines on one panel are hard to tell apart by
# colour alone, which is why the figure that follows gives each its own panel.
nine_colours <- c('#E69F00', '#56B4E9', '#009E73', '#F0E442', '#0072B2',
                  '#D55E00', '#CC79A7', '#999999', '#000000')
ggplot(filter(speculation, reference_n > 0),
       aes(year, reference_n, colour = topic)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_colour_manual(values = nine_colours) +
  labs(x = NULL, y = 'Records per year (log scale)', colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = 'right')
```

## The share of records that mention speculation

Each panel below shows one literature. The line is the yearly percentage of records that also match `speculat*`, the shaded band is a stability range computed from the yearly reference count by the Wilson method, which conveys how much the share would move with a handful of records, and the panels are ordered from the most to the least speculative literature over the whole period. The axis is cut at 15%, which clips the bands in the early years of the small literatures, where a single abstract would move the share by tens of points. Years in which a literature had no records are left blank.

```{r, fig.width = 9, fig.height = 8, fig.alt = 'Nine panels, one per literature, each showing the yearly percentage of Scopus records that mention speculation from 1980 onwards, with a shaded Wilson interval'}
wilson <- function(k, n, z = 1.96) {
  p <- k / n
  centre <- (p + z^2 / (2 * n)) / (1 + z^2 / n)
  half <- z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / (1 + z^2 / n)
  data.frame(lower = 100 * pmax(0, centre - half), upper = 100 * pmin(1, centre + half))
}

panel_data <- speculation |>
  filter(reference_n > 0) |>
  mutate(topic = factor(topic, levels = totals$topic)) |>
  bind_cols(wilson(speculation$n[speculation$reference_n > 0],
                   speculation$reference_n[speculation$reference_n > 0]))

panel_labels <- totals |>
  mutate(label = sprintf('%s\n%s records, %.1f%% with speculat*',
                         topic, format(records, big.mark = ','), share))
labeller_topic <- setNames(panel_labels$label, panel_labels$topic)

ggplot(panel_data, aes(year, percentage)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = '#0072B2', alpha = 0.15) +
  geom_line(colour = '#0072B2', linewidth = 0.7) +
  geom_point(colour = '#0072B2', size = 0.9) +
  facet_wrap(~ topic, ncol = 3, labeller = as_labeller(labeller_topic)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  coord_cartesian(ylim = c(0, 15)) +
  labs(x = NULL, y = 'Records mentioning speculation') +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(size = 10, hjust = 0),
        panel.spacing = unit(1.2, 'lines'))
```

```{r}
recent <- speculation |>
  filter(year >= 2016, year <= max(year) - 1) |>
  group_by(topic) |>
  summarise(records = sum(reference_n),
            share = 100 * sum(n) / sum(reference_n)) |>
  arrange(desc(share))
recent
```

Restricting the comparison to the last ten complete years, `r min(recent$records)` to `r format(max(recent$records), big.mark = ',')` records per literature, gives the ranking above. The literature on `r recent$topic[1]` leads with `r sprintf('%.1f', recent$share[1])`% of its records mentioning speculation, and that on `r recent$topic[nrow(recent)]` closes the list with `r sprintf('%.1f', recent$share[nrow(recent)])`%.

## What the counts can and cannot show

The measure is crude in ways that the figures make visible. A form of 'speculate' in an abstract is as often a mark of candour as of conjecture, since the authors who write 'we speculate that' are labelling a claim that others would present without a label, and a literature with a strong convention of hedged abstracts will score higher for that reason alone. The wildcard also catches 'speculative' in senses that have nothing to do with scientific conjecture, such as speculative execution in computing or speculative investment in economics, and both senses can occur in the literatures on artificial general intelligence and on the hadron collider, whose funding is discussed in economic terms. The queries define each literature by a phrase, so records that belong to a field without using its name are missed, and the literatures on language teaching and language disorders are far larger than their phrases suggest. Percentages for the early years, when a literature had a handful of records, are shown with their intervals so that a single abstract is not read as a trend.

The smallest literature, sensorimotor simulation, contains no record that mentions speculation at all, which with `r format(totals$records[totals$topic == 'sensorimotor simulation'])` records over the whole period says more about the size of the literature than about its habits.

With those limits in mind, the pattern across the language sciences is consistent with the impression that started the exercise. Literatures that theorise about unobservable or contested processes carry more explicit speculation than literatures organised around a clinical population, a classroom or an instrument, and the difference is visible in abstracts before any reading of the articles themselves. Whether the speculation is a symptom of immaturity or a healthy habit of labelling conjecture is a question that counts cannot answer. Anyone who wants to add a topic, or to see how the shares have moved, can dispatch the workflow again from the repository.

### References

Bernabeu, P., & Vogt, P. (2015). Language evolution: Current status and future directions [Paper presentation]. Tenth Language at the University of Essex (LangUE) Postgraduate Conference, Colchester, United Kingdom. https://www.researchgate.net/publication/280858062_Language_evolution_Current_status_and_future_directions

Evans, N., & Levinson, S. C. (2009). The myth of language universals: Language diversity and its importance for cognitive science. *Behavioral and Brain Sciences*, *32*(5), 429–492. https://doi.org/10.1017/S0140525X0999094X

Tomasello, M. (2009). Universal grammar is dead. *Behavioral and Brain Sciences*, *32*(5), 470–471. https://doi.org/10.1017/S0140525X09990744
