---
title: 'depictr: one visual language from the first look at the data to the audited figure'
author: ''
date: '2026-09-01'
slug: depictr-one-visual-language-from-first-look-to-final-figure
categories:
  - data visualisation
  - statistics
tags:
  - visualisation
  - linear mixed-effects models
  - Bayesian statistics
  - accessibility
  - Python
  - psycholinguistics
subtitle: ''
summary: 'The figures of an analysis are usually made with a different tool at each stage, and they show it. depictr gives one theme, one colourblind-safe palette and one set of conventions to the plots of exploration, model estimates, diagnostics, uncertainty and power, in R and in Python, and adds a check of each figure against measured accessibility thresholds. This post takes a bundled lexical-decision experiment from the first density plot to a mended, audited figure, with a mixed model, seven optimisers and a Bayesian posterior along the way.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>',
                      fig.width = 7, fig.height = 4.5)
knitr::opts_chunk$set(python.reticulate = FALSE, engine.path = list(
  python = '/tmp/claude-0/-home-user/22b7e646-ecfa-5c99-a752-a7e5438334ea/scratchpad/pyenv/bin/python'
))
options(width = 95)
xaringanExtra::use_clipboard()
parity_dir <- file.path(tempdir(), 'depictr-parity')
dir.create(parity_dir, showWarnings = FALSE)
Sys.setenv(DEPICTR_PARITY = parity_dir)
```

A paper's figures are made at different moments with different tools. The density plot from the first afternoon with the data has one look, the forest plot of the mixed model another, the diagnostics a third, and the Bayesian posterior, added at a reviewer's request, a fourth. The reader has to learn four conventions to follow one argument, and the author has to remember, at each stage, which colours are safe for readers with a colour-vision deficiency and which are not. depictr is my attempt to settle these questions once. It provides a single theme, palette and set of conventions for the plots that an analysis produces from exploration to reporting, every function returns a ggplot2 object that can be refined further, and a checking function measures the finished figure against thresholds for contrast, text size and reliance on colour. A Python package of the same name does the same over plotnine.

The data below are the experiment bundled with the package: `r length(unique(depictr::lexical_decision$participant))` participants, `r length(unique(depictr::lexical_decision$item))` items, a semantic-priming manipulation with related and unrelated primes, a visual or auditory presentation and the Zipf frequency of each word, with reaction times and accuracy. The design is what I would expect from a lexical-decision study in my own laboratory.

```{r}
library(depictr)
library(ggplot2)
library(patchwork)
packageVersion('depictr')
str(lexical_decision)
correct <- subset(lexical_decision, accuracy == 1)
```

## The first look

Column names can be given bare or quoted, and a legend can be tucked into the corner that a unimodal density leaves empty. The empirical cumulative distribution is the plot I prefer for reaction times, since it is free of bins and the quartiles can be read from it directly.

```{r, fig.alt = 'Density plot of lexical-decision reaction times by priming condition, related in blue and unrelated in orange, with the legend inside the panel'}
explore_distribution(lexical_decision, RT, group = condition, type = 'density',
                     legend_inside = TRUE,
                     title = 'Lexical-decision RT by priming condition')
```

```{r, fig.alt = 'Empirical cumulative distribution of reaction times by priming condition, with reference lines at the quartiles'}
ecdf_plot(lexical_decision, RT, group = condition,
          reference_quantiles = c(0.25, 0.5, 0.75), legend_inside = TRUE,
          title = 'Related primes are faster throughout the distribution')
```

A summary table in the format of a paper's first table, with means and standard deviations for numeric variables and counts for categorical ones, comes from the same data with one call, and an estimation plot in the style of Gardner and Altman (Ho et al., 2019) shows the raw data above the mean difference with a bootstrap interval and Hedges' g.

```{r}
summary_table(lexical_decision, vars = c('RT', 'accuracy', 'word_frequency', 'modality'),
              group = 'condition')
```

```{r, fig.height = 5.5, fig.alt = 'Estimation plot with the raw reaction times and means of the two priming conditions above and the mean difference with its bootstrap confidence interval below'}
set.seed(1)
estimation_plot(lexical_decision, RT, condition,
                title = 'RT difference: unrelated minus related priming')
```

## The model and its estimates

The analysis is a mixed model with crossed random intercepts for participants and items, fitted to the correct trials. The package tidies fitted models from lm, glm and lme4 into one table of estimates with intervals, and the forest plot reads from that table, so the same function serves a regression, a mixed model or a data frame of estimates from elsewhere.

```{r}
fit <- lmerTest::lmer(RT ~ condition + modality + word_frequency +
                        (1 | participant) + (1 | item), data = correct)
tidy_estimates(fit)
```

```{r, fig.alt = 'Forest plot of the three fixed effects of the mixed model, standardised per standard deviation of the predictor, with 95% confidence intervals'}
term_labels <- c(conditionunrelated = 'Unrelated priming',
                 modalityauditory   = 'Auditory modality',
                 word_frequency     = 'Word frequency (Zipf)')
coefficient_plot(fit, standardise = TRUE, order = 'ascending', labels = term_labels,
                 title = 'Effects on RT per SD of each predictor')
```

Standardising by the predictor's standard deviation puts a slope per Zipf unit and a contrast between two categories on one comparable axis, and the axis label records the convention. Predicted values come from the fixed effects with a band from the design matrix and the covariance of the estimates, and the conditional modes of the random effects are drawn as a caterpillar plot, one panel per grouping factor.

```{r, fig.alt = 'Predicted reaction time across word frequency with a confidence band'}
effects_plot(fit, 'word_frequency', title = 'Predicted RT across word frequency')
```

```{r, fig.height = 6, fig.alt = 'Caterpillar plots of the by-participant and by-item random intercepts, sorted, with conditional intervals'}
random_effects_plot(fit, title = 'By-participant and by-item departures')
```

Two specifications can be overlaid on one forest plot to show whether dropping a predictor moves the other estimates, and a fit table collects the usual comparison statistics.

```{r, fig.alt = 'Forest plot comparing the estimates of the full model and a reduced model without modality'}
reduced <- lmerTest::lmer(RT ~ condition + word_frequency +
                            (1 | participant) + (1 | item), data = correct)
compare_models(Reduced = reduced, Full = fit, order = 'descending',
               labels = term_labels,
               title = 'Does dropping modality move the other effects?')
model_fit_table(Reduced = reduced, Full = fit)
```

## Convergence, a Bayesian check and power

A convergence warning from lme4 is best answered by refitting with every optimiser and comparing the estimates, a practice I have [written about before](/2023/a-new-function-to-plot-convergence-diagnostics-from-lme4-allfit/). depictr's plot for that comparison takes a tidy data frame of optimiser, term and value, and the package ships the summary of an `allFit()` run on this very model, so the figure is drawn without the refits.

```{r, fig.alt = 'Dot plot of each fixed effect as estimated by seven optimisers, one panel per term, showing that the estimates coincide'}
af <- readRDS(system.file('extdata', 'allfit_lexdec.rds', package = 'depictr'))
fx <- af$fixef
opt_long <- data.frame(optimizer = rep(rownames(fx), times = ncol(fx)),
                       term = rep(colnames(fx), each = nrow(fx)),
                       value = as.vector(fx))
rownames(fx)
optimizer_fixef_plot(opt_long, labels = term_labels,
                     title = 'Fixed effects across seven optimisers')
```

A Bayesian version of the same model answers the next question. The package bundles a thousand draws from the posterior of the same model fitted with brms, and the plot overlays the frequentist estimate on the full posterior of each term, with a note that the frequentist analysis had no prior. The posterior can also be drawn on its own, with a region of practical equivalence and the probability of direction annotated per parameter (Kruschke, 2018).

```{r, fig.height = 5, fig.alt = 'Half-eye plots of the posterior of each fixed effect with the frequentist point estimate and interval overlaid'}
draws <- readRDS(system.file('extdata', 'lexdec_draws.rds', package = 'depictr'))
dim(draws)
frequentist_bayesian_plot(fit, draws, intercept = FALSE,
                          note_frequentist_no_prior = TRUE, labels = term_labels,
                          title = 'Frequentist estimate over the Bayesian posterior')
```

```{r, fig.alt = 'Posterior distributions of the three slopes with a region of practical equivalence of plus or minus five milliseconds and the probability of direction'}
slopes <- draws[c('conditionunrelated', 'modalityauditory', 'word_frequency')]
posterior_plot(slopes, style = 'halfeye', rope = c(-5, 5), pd = TRUE,
               labels = term_labels,
               title = 'Fixed-effect posteriors (ms), with ROPE and pd')
```

Whether the study was large enough is the third question. The package ships the summary of a simr power curve for the word-frequency slope at five sample sizes, and the plotting function accepts such a summary as a data frame, so simr is not needed to draw it. A fuller treatment of design analysis by simulation is the subject of the [pilotr post](/2026/pilotr-pilot-the-study-before-running-it/).

```{r, fig.alt = 'Power curve for the word-frequency effect at 12 to 60 participants with confidence intervals'}
pc <- readRDS(system.file('extdata', 'powercurve_lexdec_summary.rds', package = 'depictr'))
pc
power_curve_plot(pc, x_lab = 'Number of participants',
                 title = 'Power for the word-frequency effect')
```

## Diagnostics that match the model class

The diagnostics function recognises the class of the model. For a binomial model of accuracy, it draws binned residuals (Gelman & Hill, 2007) and a quantile-quantile plot of randomised quantile residuals (Dunn & Smyth, 1996), with a seed so that the randomisation is reproducible, in place of the panels that suit a linear model.

```{r, fig.height = 6, fig.alt = 'Four diagnostic panels for a binomial model of accuracy, including binned residuals and a quantile-quantile plot of randomised quantile residuals'}
gfit <- glm(accuracy ~ word_frequency + condition + modality,
            data = lexical_decision, family = binomial)
residual_diagnostics_plot(gfit, seed = 1, title = 'Accuracy model: GLM diagnostics')
```

## The audit

The palette is based on the eight colours of Okabe and Ito (2008), with the package's own blue leading and a grey in place of black, and the package checks it the way it checks any palette, by the CIE76 colour difference between every pair under normal vision and under each dichromacy simulated at full severity with the model of Machado et al. (2009). Palette safety does not settle figure safety, so a second function inspects a built plot and reports nine measurements beside their thresholds.

```{r}
depictr_palette()
palette_safety()
```

```{r, fig.height = 2.6, fig.alt = 'Swatches of the eight palette colours as a person with deuteranopia sees them, labelled with the original hexadecimal codes'}
palette_preview(cvd = 'deutan')
```

```{r}
p_density <- explore_distribution(lexical_decision, RT, group = condition,
                                  type = 'density', legend_inside = TRUE)
check_figure(p_density)
```

Two rows fail on the default two-colour density. The orange sits at a contrast of about 2.25 to 1 against white, where the guideline for graphical objects asks for 3 to 1, and colour alone distinguishes the conditions. Both are corrected in one line, by swapping in the darker vermillion and adding line type as a redundant channel, since the plot is an ordinary ggplot2 object. Text drawn for a 17.8 cm figure arrives at half its size in an 8.9 cm journal column, and the check can be run at that width.

```{r, fig.alt = 'Empirical cumulative distributions by priming condition in blue and vermillion with different line types, the mended version of the earlier figure'}
p_mended <- ecdf_plot(lexical_decision, RT, group = condition,
                      palette = c('#005b96', '#d55e00'), legend_inside = TRUE,
                      title = 'Corrected: darker vermillion, with line type as a second channel') +
  aes(linetype = condition)
check_figure(p_mended)[, c('check', 'measured', 'threshold', 'verdict')]
subset(check_figure(p_density, width_cm = 8.9), check == 'text_size')
p_mended
```

The documentation states the limit of the guarantee beside the claim. The eight colours pass every colour-vision check, and the closest pair in greyscale, the orange and the sky blue, differ by about one unit of CIE lightness, so a figure that may be printed in black and white needs a second channel as well.

## Composition and the Python twin

Panels are composed with duplicate legends collected, an overall title and tags, and the package options set the base size and the accent colour once for every theme, scale and plot.

```{r, fig.width = 11, fig.height = 8, fig.alt = 'Four-panel composite with the density plot, the mixed-model forest plot, the posterior with the frequentist overlay and the mended cumulative distribution'}
p_coef <- coefficient_plot(fit, order = 'ascending', labels = term_labels)
p_fbp <- frequentist_bayesian_plot(fit, draws, intercept = FALSE, labels = term_labels)
arrange_plots(p_density + labs(title = 'A first look'),
              p_coef    + labs(title = 'Mixed-model estimates'),
              p_fbp     + labs(title = 'Posterior with frequentist overlay'),
              p_mended  + labs(title = 'Audited and mended'),
              ncol = 2, title = 'One visual language, from exploration to audit',
              tag_levels = 'A')
```

The Python package uses the same function names and the same thresholds. Below, the R session writes the data, the density audit and the colour-vision simulation to disk. The Python session then reads the data, repeats the audit and the simulation, and compares the two sets of results. The mixed model is the one place where the engines differ by construction, lme4 against statsmodels, so the comparison below uses an ordinary regression on the correct trials, which both fit identically once the reference level of the categorical predictor is set to match.

```{r}
write.csv(lexical_decision, file.path(parity_dir, 'lexical_decision.csv'), row.names = FALSE)
write.csv(check_figure(p_density), file.path(parity_dir, 'audit_R.csv'), row.names = FALSE)
writeLines(simulate_cvd(depictr_palette(), 'deutan'), file.path(parity_dir, 'deutan_R.txt'))
ols <- lm(RT ~ condition + modality + word_frequency, data = correct)
write.csv(tidy_estimates(ols), file.path(parity_dir, 'ols_R.csv'), row.names = FALSE)
```

```{python}
import os
import numpy as np, pandas as pd
import statsmodels.formula.api as smf
import depictr as dp

parity = os.environ['DEPICTR_PARITY']
ld = pd.read_csv(os.path.join(parity, 'lexical_decision.csv'))
correct = ld[ld['accuracy'] == 1]

audit_py = dp.check_figure(dp.explore_distribution(
    ld, 'RT', group='condition', kind='density', legend_inside=True))
audit_r = pd.read_csv(os.path.join(parity, 'audit_R.csv'))
deutan_r = open(os.path.join(parity, 'deutan_R.txt')).read().split()
# R's factor keeps the levels of `modality` in the order 'visual', 'auditory';
# pandas reads the column as text and patsy would sort it alphabetically, which
# would silently flip the sign of that contrast. The reference is set explicitly.
formula = ("RT ~ condition + C(modality, Treatment(reference='visual')) "
           "+ word_frequency")
ols_py = dp.tidy_estimates(smf.ols(formula, data=correct).fit())
ols_r = pd.read_csv(os.path.join(parity, 'ols_R.csv'))

print('depictr', dp.__version__)
print('audit verdicts identical:', audit_py['verdict'].tolist() == audit_r['verdict'].tolist())
print('audit measurements within 0.005:',
      np.allclose(audit_py['measured'].astype(float), audit_r['measured'].astype(float), atol=0.005))
print('deuteranopia simulation identical:',
      dp.simulate_cvd(dp.depictr_palette(), 'deutan') == deutan_r)
print('OLS estimates and intervals within 1e-6:',
      np.allclose(ols_py[['estimate', 'conf_low', 'conf_high']].to_numpy(),
                  ols_r[['estimate', 'conf.low', 'conf.high']].to_numpy(), atol=1e-6))
```

The package does not fit models, so the choice of model, contrasts and priors remains the analyst's, and the plots show what was fitted. The colour-difference thresholds are conventions, and the audit reports measurements against them so that a different threshold can be applied. Beyond eight categories, the palette interpolates and warns, and the guarantee lapses, which is one more reason to prefer facets to colour when a figure has many groups.

### References

Dunn, P. K., & Smyth, G. K. (1996). Randomized quantile residuals. *Journal of Computational and Graphical Statistics*, *5*(3), 236–244. https://doi.org/10.1080/10618600.1996.10474708

Gelman, A., & Hill, J. (2007). *Data analysis using regression and multilevel/hierarchical models*. Cambridge University Press. https://doi.org/10.1017/CBO9780511790942

Kruschke, J. K. (2018). Rejecting or accepting parameter values in Bayesian estimation. *Advances in Methods and Practices in Psychological Science*, *1*(2), 270–280. https://doi.org/10.1177/2515245918771304

Ho, J., Tumkaya, T., Aryal, S., Choi, H., & Claridge-Chang, A. (2019). Moving beyond P values: Data analysis with estimation graphics. *Nature Methods*, *16*(7), 565–566. https://doi.org/10.1038/s41592-019-0470-3

Machado, G. M., Oliveira, M. M., & Fernandes, L. A. F. (2009). A physiologically-based model for simulation of color vision deficiency. *IEEE Transactions on Visualization and Computer Graphics*, *15*(6), 1291–1298. https://doi.org/10.1109/TVCG.2009.113

Okabe, M., & Ito, K. (2008). *Color universal design (CUD): How to make figures and presentations that are friendly to colorblind people*. https://jfly.uni-koeln.de/color/ (Original work published 2002)
