---
title: 'A new function to plot convergence diagnostics from lme4::allFit()'
author: R package build
date: ~
slug: a-new-function-to-plot-convergence-diagnostics-from-lme4-allfit
categories:
  - R
  - linear mixed-effects models
tags:
  - R
  - linear mixed-effects models
  - data visualisation
  - statistics
subtitle: ''
summary: 'When a model has struggled to find enough information in the data to account for every predictor---especially for every random effect---, convergence warnings appear (Brauer & Curtin, 2018; Singmann & Kellen, 2019). In this article, I review the issue of convergence before presenting a new plotting function in R that facilitates the visualisation of the fixed effects fitted by different optimization algorithms (also dubbed optimizers).'
authors: []
lastmod: '2021-10-20T18:46:08+01:00'
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


```{r setup, include = FALSE}

# Basic knitr options
knitr::opts_chunk$set(message = FALSE, warning = TRUE, error = TRUE, echo = TRUE, fig.align = 'center')

# Enable copy of code blocks to clipboard
library(xaringanExtra)
xaringanExtra::use_clipboard()

```


Linear mixed-effects models (LMM) offer a consistent way of performing regression and analysis of variance tests which allows accounting for non-independence in the data. Over the past decades, LMMs have subsumed most of the General Linear Model, with a stead increase in popularity (Meteyard & Davies, 2020). Since their conception, LMMs have presented the challenge of model *convergence*. In essence, the issue of convergence boils down to the widespread tension between parsimony and completeness in data analysis. That is, on the one hand, a good model must allow an accurate, parsimonious analysis of each predictor, and thus, it must not be overfitted with too many parameters. Yet, on the other hand, the model must be complete enough to account for a sufficient amount of variation in the data. The latter requirement can lead to convergence problems when a model struggles to find enough information in the data to account for every predictor---especially for every random effect. In such a circumstance, convergence warnings appear (Brauer & Curtin, 2018; Singmann & Kellen, 2019). In this article, I review the issue of convergence before presenting a new plotting function in R that facilitates the diagnosis of convergence by visualising the fixed effects fitted by different optimization algorithms (also dubbed optimizers).

Both fixed and random effects comprise intercepts and slopes. The pressure exerted by each of those types of effects on the model is determined by the number of data points involved by each. First, slopes are more demanding than intercepts, as they involve a (far) larger number of data points. Second, random effects are more demanding than fixed effects, as the former entail the number of estimates required for fixed effects *times* the number of levels in the grouping factor. As a result, on the most lenient end of the scale lies the fixed intercept, and on the heaviest end lie the random slopes. Convergence warnings in LMMs are often due to the random slopes alone. 

Sounds easy, then! Not inviting the random slopes to the party should solve the problem. Indeed, since random slopes involve the highest number of estimates by far, removing them does often remove convergence warnings. This, however, leads to a different problem. Surrendering the information provided by random slopes can result in the violation of the assumption of independence. For years, the removal of random slopes due to convergence warnings was standard practice. Currently, in contrast, proposals increasingly consider other options, such as removing random effects if they do not significantly improve the fit of the model (Matuschek et al., 2017), and keeping the random slopes in spite of the convergence warnings (Brauer & Curtin, 2018; Singmann & Kellen, 2019).

Tying in with the latter option is a proposal by the developers of the 'lme4' package that uses a part of the 'lme4' *engine* called optimizer. Each model can have one optimizer out of a range of options. The seven widely-available optimizers are:

- Bobyqa
- Nelder-Mead
- nlminbwrap
- nmkbw
- optimx.L-BFGS-B
- nloptwrap.NLOPT_LN_NELDERMEAD
- nloptwrap.NLOPT_LN_BOBYQA


To assess whether convergence warnings render the results invalid, or on the contrary, the results can be deemed valid in spite of the warnings, Bates et al. (2021) suggest refitting models affected by convergence warnings with a variety of optimizers. The authors argue that if the different optimizers produce practically-equivalent results, the results are valid.

The `allFit` function from the 'lme4' package allows the refitting of models using a number of optimizers. To use the seven optimizers listed above, two extra packages must be installed: 'dfoptim' and 'optimx' (see [lme4 manual](https://cran.r-project.org/web/packages/lme4/lme4.pdf)). The output from `allFit()` contains several statistics on the fixed and the random effects fitted by each optimizer (see [example](https://github.com/lme4/lme4/issues/512#issue-425198940)). 

Several R users have ventured into [plotting this output](https://www.google.com/search?q=%22ggplot%22+%22allfit%22+optimizers) but there is not a function in 'lme4' yet at the time of writing (Oct 2021). I have just developed a function that takes the output from `allFit()`, tidies it, selects the fixed effects and plots them using 'ggplot2'. The function might be integrated in the 'lme4' package in the near future, but for now it is available below.

Below are the optional arguments allowed by the function, with their default values.

```

# Set the same Y axis limits in every plot
shared_y_axis_limits = TRUE,

# Multiply Y axis limits by a factor (only 
# available if shared_y_axis_limits = TRUE)
multiply_y_axis_limits = 1, 

# Number of decimal points
decimal_points = NULL,

# Select predictors
select_predictors = NULL, 

# Number of rows
nrow = NULL, 

# Y axis title
y_title = 'Fixed effect',

# Alignment of the Y axis title
y_title_hjust = NULL,

# Add number to the names of optimizers
number_optimizers = TRUE,

# Replace colon in interactions with x
interaction_symbol_x = TRUE

```


The argument `shared_y_axis_limits` deserves a comment. It allows using the same Y axis limits (i.e., range) in all plots or using plot-specific limits. It is `TRUE` by default to prevent overinterpretations of small differences across optimizers. In contrast, when `shared_y_axis_limits = FALSE`, plot-specific limits are used, which results in a narrower range of values in the Y axis. Since data points will span the entire Y axis in that case, any difference across optimizers---regardless of its relative importance---might be perceived as large, unless the specific range of values in each plot is noticed.


## The function and its use

Below is the new `plot.fixef.allFit` function (of course it could be given any other name). The function can be copied by clicking on the `Copy Code` button at the top right corner. 

<div style = 'height: 800px; border: 0.5px dotted grey; padding: 10px; resize: both; overflow: auto;'>

```{r}

# Plot the results from the fixed effects produced by different optimizers. This function 
# takes the output from lme4::allFit(), tidies it, selects fixed effects and plots them.

plot.fixef.allFit = function(allFit_output, 
                             # Set the same Y axis limits in every plot
                             shared_y_axis_limits = TRUE,
                             # Multiply Y axis limits by a factor (only 
                             # available if shared_y_axis_limits = TRUE)
                             multiply_y_axis_limits = 1, 
                             # Number of decimal points
                             decimal_points = NULL,
                             # Select predictors
                             select_predictors = NULL, 
                             # Number of rows
                             nrow = NULL, 
                             # Y axis title
                             y_title = 'Fixed effect',
                             # Alignment of the Y axis title
                             y_title_hjust = NULL,
                             # Add number to the names of optimizers
                             number_optimizers = TRUE,
                             # Replace colon in interactions with x
                             interaction_symbol_x = TRUE) {
  
  # Data wrangling
  if (!requireNamespace('dplyr')) install.packages('dplyr')
  if (!requireNamespace('reshape2')) install.packages('reshape2')
  # Text processing
  if (!requireNamespace('stringr')) install.packages('stringr')
  # Set number of decimal points
  if (!requireNamespace('scales')) install.packages('scales')
  # Plotting
  if (!requireNamespace('ggplot2')) install.packages('ggplot2')
  # Matrix of plots
  if (!requireNamespace('patchwork')) install.packages('patchwork')
  
  require(dplyr)
  require(reshape2)
  require(stringr)
  require(scales)
  require(ggplot2)
  require(patchwork)
  
  # Tidy allFit output
  
  # Extract fixed effects from the allFit() output
  allFit_fixef = summary(allFit_output)$fixef %>%  # Select fixed effects in the allFit results
    reshape2::melt() %>%  # Structure the output as a data frame
    rename('Optimizer' = 'Var1', 'fixed_effect' = 'Var2')  # set informative names
  
  # If number_optimizers = TRUE, assign number to each optimizer and place it before its name
  if(number_optimizers == TRUE) {
    allFit_fixef$Optimizer = paste0(as.numeric(allFit_fixef$Optimizer), '. ', allFit_fixef$Optimizer)
  }
  
  # If select_predictors were specified, select them along with the intercept (the latter required)
  if(!is.null(select_predictors)) {
    allFit_fixef = allFit_fixef %>% filter(fixed_effect %in% c('(Intercept)', select_predictors))
  }
  
  # Order variables
  allFit_fixef = allFit_fixef[, c('Optimizer', 'fixed_effect', 'value')]
  
  # PLOT. The overall plot is formed of a first row containing the intercept and the legend 
  # (intercept_plot), and a second row containing the predictors (predictors_plot), 
  # which may in turn occupy several rows.
  
  # If multiply_y_axis_limits was specified but shared_y_axis_limits = FALSE,
  # warn that shared_y_axis_limits is required.
  if(!multiply_y_axis_limits == 1 & shared_y_axis_limits == FALSE) {
    message('The argument `multiply_y_axis_limits` has not been used because it requires `shared_y_axis_limits` set to TRUE.')
  }
  
  # If extreme values were entered in y_title_hjust, show warning
  if(!is.null(y_title_hjust)) {
    if(y_title_hjust < 0.5 | y_title_hjust > 6) {
      message('NOTE: For y_title_hjust, a working range of values is between 0.6 and 6.')
    }
  }
  
  # If decimal_points were specified, convert number to the format used in 'scales' package
  if(!is.null(decimal_points)) {
    decimal_points = 
      ifelse(decimal_points == 1, 0.1, 
             ifelse(decimal_points == 2, 0.01, 
                    ifelse(decimal_points == 3, 0.001, 
                           ifelse(decimal_points == 4, 0.0001, 
                                  ifelse(decimal_points == 5, 0.00001, 
                                         ifelse(decimal_points == 6, 0.000001, 
                                                ifelse(decimal_points == 7, 0.0000001, 
                                                       ifelse(decimal_points == 8, 0.00000001, 
                                                              ifelse(decimal_points == 9, 0.000000001, 
                                                                     ifelse(decimal_points == 10, 0.0000000001,
                                                                            ifelse(decimal_points == 11, 0.00000000001,
                                                                                   ifelse(decimal_points == 12, 0.000000000001,
                                                                                          ifelse(decimal_points == 13, 0.0000000000001,
                                                                                                 ifelse(decimal_points == 14, 0.00000000000001,
                                                                                                        ifelse(decimal_points > 15, 0.000000000000001, 
                                                                                                               0.001
                                                                                                        )))))))))))))))
  }
  
  # First row: intercept_plot
  
  # Select intercept data only
  intercept = allFit_fixef %>% filter(fixed_effect == '(Intercept)')
  
  intercept_plot = intercept %>%
    ggplot(., aes(fixed_effect, value, colour = Optimizer)) +
    geom_point(position = position_dodge(1)) +
    facet_wrap(~fixed_effect, scale = 'free') +
    guides(colour = guide_legend(title.position = 'left')) +
    theme_bw() + theme(axis.title = element_blank(), axis.ticks.x = element_blank(),
                       axis.text.x = element_blank(), 
                       strip.text = element_text(size = 10, margin = margin(t = 4, b = 6)),
                       strip.background = element_rect(fill = 'grey96'),
                       legend.margin = margin(0.3, 0, 0.8, 1, 'cm'), 
                       legend.title = element_text(size = unit(15, 'pt'), angle = 90, hjust = 0.5))
  
  # Second row: predictors_plot
  
  # Select all predictors except intercept
  predictors = allFit_fixef %>% filter(!fixed_effect == '(Intercept)')
  
  # If interaction_symbol_x = TRUE (default), replace colon with times symbol x between spaces
  if(interaction_symbol_x == TRUE) {
    # Replace colon in interactions with \u00D7, i.e., x; then set factor class
    predictors$fixed_effect = predictors$fixed_effect %>% str_replace_all(':', ' \u00D7 ') %>% factor()
  }
  
  # Order predictors as in the original output from lme4::allFit()
  predictors$fixed_effect = factor(predictors$fixed_effect, levels = unique(predictors$fixed_effect))
  
  # Set number of rows for the predictors excluding the intercept.
  # First, if nrow argument specified, use it
  if(!is.null(nrow)) {
    predictors_plot_nrow = nrow - 1  # Subtract 1 as intercept row not considered
    
    # Else, if nrow argument not specified, calculate sensible number of rows: i.e., divide number of
    # predictors (exc. intercept) by 2 and round up the result. For instance, 7 predictors --> 3 rows
  } else predictors_plot_nrow = (length(unique(predictors$fixed_effect)) / 2) %>% ceiling()
  
  predictors_plot = ggplot(predictors, aes(fixed_effect, value, colour = Optimizer)) +
    geom_point(position = position_dodge(1)) +
    facet_wrap(~fixed_effect, scale = 'free',
               # Note that predictors_plot_nrow was defined a few lines above
               nrow = predictors_plot_nrow) +
    labs(y = y_title) +
    theme_bw() + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                       axis.ticks.x = element_blank(),
                       axis.title.y = element_text(size = 14, margin = margin(0, 15, 0, 5, 'pt')),
                       strip.text = element_text(size = 10, margin = margin(t = 4, b = 6)),
                       strip.background = element_rect(fill = 'grey96'), legend.position = 'none')
  
  # Below, the function scale_y_continuous is applied conditionally to avoid overriding settings. First, 
  # if shared_y_axis_limits = TRUE and decimal_points were specified, set the same Y axis limits in 
  # every plot and set decimal_points. By default, also expand limits by a seventh of its original limit, 
  # and allow further multiplication of limits through multiply_y_axis_limits.
  if(shared_y_axis_limits == TRUE & !is.null(decimal_points)) {
    
    intercept_plot = intercept_plot +
      scale_y_continuous(limits = c(min(allFit_fixef$value) - allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits,
                                    max(allFit_fixef$value) + allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits), 
                         # Set number of decimal points
                         labels = scales::label_number(accuracy = decimal_points))
    
    predictors_plot = predictors_plot + 
      scale_y_continuous(limits = c(min(allFit_fixef$value) - allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits,
                                    max(allFit_fixef$value) + allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits), 
                         # Set number of decimal points
                         labels = scales::label_number(accuracy = decimal_points))
    
    # Else, if shared_y_axis_limits = TRUE but decimal_points were not specified, do as above but without
    # setting decimal_points.
  } else if(shared_y_axis_limits == TRUE & is.null(decimal_points)) {
    
    intercept_plot = intercept_plot +
      scale_y_continuous(limits = c(min(allFit_fixef$value) - allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits,
                                    max(allFit_fixef$value) + allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits))
    
    predictors_plot = predictors_plot + 
      scale_y_continuous(limits = c(min(allFit_fixef$value) - allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits,
                                    max(allFit_fixef$value) + allFit_fixef$value %>% abs %>% max / 7 * multiply_y_axis_limits))
    
    # Else, if shared_y_axis_limits = FALSE and decimal_points were specified, set decimal_points. 
  } else if(shared_y_axis_limits == FALSE & !is.null(decimal_points)) {
    
    intercept_plot = intercept_plot +
      scale_y_continuous(labels = scales::label_number(accuracy = decimal_points))
    
    predictors_plot = predictors_plot + 
      scale_y_continuous(labels = scales::label_number(accuracy = decimal_points))
  }
  
  # Plot matrix: based on number of predictors_plot_nrow, adjust height of Y axis title
  # (unless specified by user), and assign space to intercept_plot and predictors_plot
  if(predictors_plot_nrow == 1) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 3.6))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 11, l = 0)      # predictors row(s)
    )
    
  } else if(predictors_plot_nrow == 2) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 1.4))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 16, l = 0)      # predictors row(s)
    )
    
  } else if(predictors_plot_nrow == 3) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 0.92))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 21, l = 0)      # predictors row(s)
    )
    
  } else if(predictors_plot_nrow == 4) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 0.8))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 26, l = 0)      # predictors row(s)
    )
    
  } else if(predictors_plot_nrow == 5) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 0.73))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 31, l = 0)      # predictors row(s)
    )
    
  } else if(predictors_plot_nrow > 5) {
    
    # If y_title_hjust specified by user, use it
    if(!is.null(y_title_hjust)) {
      predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = y_title_hjust))
      # Otherwise, set a sensible height
    } else predictors_plot = predictors_plot + theme(axis.title.y = element_text(hjust = 0.65))
    
    layout = c(
      area(t = 1.5, r = 8.9, b = 6.8, l = 0),  # intercept row
      area(t = 7.3, r = 9, b = 36, l = 0)      # predictors row(s)
    )
    
    # Also, advise user to consider distributing predictors into several plots
    message('Many rows! Consider distributing predictors into several plots using argument `select_predictors`')
  } 
  
  # Return matrix of plots
  wrap_plots(intercept_plot, predictors_plot, design = layout,
             # The 2 below corresponds to intercept_plot and predictors_plot
             nrow = 2)
  
}

```

</div>

<br>

Let's test the function on a new analysis of the English Lexicon Project (Balota et al., 2007; Yap et al., 2012) that I've conducted for a forthcoming study.

```{r}

# Read in allFit() output
m1_allFit_convergence = readRDS('m1_allFit_convergence.rds')

# To select specific predictors, first return their names
colnames(summary(m1_allFit_convergence)$fixef)

```


Let's plot a subset of the effects using the argument `select_predictors`. Notice that the intercept is always plotted on the first row, alongside the legend listing the optimizers.

```{r demo-plot.fixef.allFit-function-1, fig.dim = c(9, 7)}

plot.fixef.allFit(m1_allFit_convergence, 
                  
                  select_predictors = c("z_vocabulary_age",
                                        "z_recoded_participant_gender",
                                        "z_word_frequency",
                                        "z_vocabulary_age:z_word_frequency",
                                        "z_recoded_participant_gender:z_word_frequency"), 
                  
                  # Increase padding at top and bottom of Y axis
                  multiply_y_axis_limits = 1.3)

```


The plot produced by `plot.fixef.allFit()` by default replaces the colons in interaction effects (e.g., `z_vocabulary_age:z_word_frequency`) with ' &times; ' to facilitate the visibility (otherwise use `interaction_symbol_x = FALSE`). Yet, it is important to note that any interactions passed to `select_predictors` must have the colon, as that is the symbol present in the `lme4::allFit()` output.

The output of the function is a [ggplot2](https://ggplot2.tidyverse.org/) object, which can be stored for further use as in the following example. New parameters are also demonstrated below; specifically, the Y axis is unique to each plot (as `shared_y_axis_limits = FALSE`), 7 decimal points are shown, and the Y axis title is specified.

```{r demo-plot.fixef.allFit-function-2, fig.dim = c(9, 4)}

plot_m1_allFit_convergence = 
  
  plot.fixef.allFit(m1_allFit_convergence, 
                    
                    select_predictors = c("z_vocabulary_age",
                                          "z_vocabulary_age:z_word_frequency"), 
                    
                    # Use plot-specific Y axis limits
                    shared_y_axis_limits = FALSE,
                    
                    decimal_points = 7, 
                    
                    # Move up Y axis title
                    y_title_hjust = 5.5,
                    
                    y_title = 'Fixed effect (\u03B2)')  # \u03B2 = beta letter

# Modify aspect further using `ggplot2::theme()`
plot_m1_allFit_convergence =
  plot_m1_allFit_convergence + theme(axis.title.y = element_text(size = 12))

# Print plot
plot_m1_allFit_convergence

# Plot can be saved to disk as pdf, png, etc. through `ggplot2::ggsave()`
# ggsave('plot_m1_allFit_convergence.pdf', plot_m1_allFit_convergence, 
#        device = cairo_pdf, width = 9, height = 9, dpi = 900)

```


<br>

## References

<div style = "text-indent:-2em; margin-left:2em;">

Balota, D. A., Yap, M. J., Hutchison, K. A., Cortese, M. J., Kessler, B., Loftis, B., Neely, J. H., Nelson, D. L., Simpson, G. B., & Treiman, R. (2007). The English Lexicon Project. *Behavior Research Methods, 39*, 445–459. https://doi.org/10.3758/BF03193014

Bates, D., Maechler, M., Bolker, B., Walker, S., Christensen, R. H. B., Singmann, H., Dai, B., Scheipl, F., Grothendieck, G., Green, P., Fox, J., Bauer, A., & Krivitsky, P. N. (2021). *Package 'lme4'.* CRAN. https://cran.r-project.org/web/packages/lme4/lme4.pdf

Brauer, M., & Curtin, J. J. (2018). Linear mixed-effects models and the analysis of nonindependent data: A unified framework to analyze categorical and continuous independent variables that vary within-subjects and/or within-items. *Psychological Methods, 23*(3), 389–411. https://doi.org/10.1037/met0000159

Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing type 1 error and power in linear mixed models. *Journal of Memory and Language, 94*, 305–315. https://doi.org/10.1016/j.jml.2017.01.001

Meteyard, L., & Davies, R. A. (2020). Best practice guidance for linear mixed-effects models in psychological science. *Journal of Memory and Language, 112*, 104092. https://doi.org/10.1016/j.jml.2020.104092

Singmann, H., & Kellen, D. (2019). An introduction to mixed models for experimental psychology. In D. H. Spieler & E. Schumacher (Eds.), New methods in cognitive psychology (pp. 4–31). Psychology Press.

Yap, M. J., Balota, D. A., Sibley, D. E., & Ratcliff, R. (2012). Individual differences in visual word recognition: Insights from the English Lexicon Project. *Journal of Experimental Psychology: Human Perception and Performance, 38*, 1, 53–79. https://doi.org/10.1037/a0024177

</div>


