---
title: 'lexsync: From a word-frequency corpus to an EEG-ready experiment'
author: ''
date: '2026-09-01'
apa_captions: true
slug: lexsync-from-corpus-to-eeg-ready-experiment
code_folding: hide
categories:
  - R
  - psycholinguistics
  - research methods
tags:
  - s
  - R
  - psycholinguistics
  - experimental stimuli
  - electroencephalography
  - reproducibility
  - Python
  - PsychoPy
  - OpenSesame
  - jsPsych
subtitle: ''
summary: 'Selecting words is only the first step in preparing a psycholinguistic experiment. This post uses lexsync to match high- and low-frequency items, balance them across lists, and export timed PsychoPy and OpenSesame scripts that send electroencephalography (EEG) triggers, plus a jsPsych version that logs event markers.'
abstract: 'This tutorial shows how lexsync turns a lexical design into checked stimulus lists and experiment files. It separates candidate selection from matching, evaluates balance on the intended controls, and carries the resulting item table into PsychoPy and OpenSesame, which send electroencephalography (EEG) triggers, and into jsPsych, which records an event marker with every trial.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
package_docs:
  - name: R documentation
    url: https://pablobernabeu.github.io/lexsync/r/
  - name: Python documentation
    url: https://pablobernabeu.github.io/lexsync/python/
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 92)
xaringanExtra::use_clipboard()
output_dir <- file.path(tempdir(), 'lexsync-blog')
unlink(output_dir, recursive = TRUE)
dir.create(output_dir, showWarnings = FALSE)
```

A high- versus low-frequency word experiment sounds simple until the materials are assembled. The frequency bands must differ, while length and orthographic similarity should not. The selected words then need to be divided across presentation lists without creating a new imbalance. Finally, the same item table must reach the experiment software with its trial order, timing and electroencephalography (EEG) triggers intact.

I wrote [lexsync](https://github.com/pablobernabeu/lexsync) (Bernabeu, 2026) to keep those stages in one reproducible workflow. The package stores the design separately from the candidate corpus, reports the balance that was achieved, and exports runnable experiments for PsychoPy, OpenSesame and jsPsych. Its R and Python implementations read the same design format, and the examples below use R.

## Development

The workflow below keeps each design decision in view as it defines the candidate pool, checks the realised balance and passes the same specification on to the presentation software. Every step works from the design, so a change to it reaches the generated experiment without being retyped.

```{r}
library(lexsync)
library(ggplot2)
packageVersion('lexsync')

schema_path <- system.file('extdata', 'schema.yaml', package = 'lexsync')
lexicon_path <- system.file('extdata', 'en_example.csv', package = 'lexsync')
schema <- yaml::read_yaml(schema_path)
lexicon <- load_lexicon(lexicon_path, schema, language = 'english')
```

### Put the Design in Data

The design below asks for 40 words in each frequency condition, matched on length, neighbourhood density and orthographic Levenshtein distance 20 (OLD20). Those three controls and frequency itself are then balanced across two presentation lists. Frequency is expressed on the Zipf scale (van Heuven et al., 2014), and OLD20 measures orthographic similarity as the mean edit distance between a word and its 20 nearest neighbours (Yarkoni et al., 2008).

```{r}
design <- list(
  name = 'frequency_eeg',
  language = 'english',
  n_per_condition = 40,
  pool_filters = list(length = c(3, 7), frequency = c(3.8, 7.0)),
  conditions = list(
    list(name = 'high_frequency', define_by = list(frequency = c(5.2, 7.0))),
    list(name = 'low_frequency', define_by = list(frequency = c(3.8, 4.4)))
  ),
  match_on = c('length', 'n_density', 'old20'),
  counterbalance = list(
    lists = 2,
    optimise = TRUE,
    balance_on = c('length', 'n_density', 'old20', 'frequency')
  )
)

pool <- build_pool(lexicon, design$pool_filters)
candidate_counts <- vapply(
  design$conditions,
  function(condition) nrow(build_pool(pool, condition$define_by)),
  integer(1)
)
names(candidate_counts) <- vapply(design$conditions, `[[`, character(1), 'name')
candidate_counts
```

The high-frequency band is the smaller candidate set, so it limits both the number and the quality of possible matches. Counting the candidates before matching shows whether a later shortfall stems from the size of this pool.

### Select and Check the Words

`match_stimuli()` spreads anchor words evenly across the first condition in the design, here the smaller high-frequency band, and finds nearby words from the other condition in the standardised space of the control variables. Standardising first means that a given distance counts the same whichever control it falls on.

```{r}
stimuli <- match_stimuli(pool, design, schema)
head(stimuli[, c('set', 'condition', 'word', 'frequency', 'length',
                  'n_density', 'old20')])
```

The report treats matching as a claim of equivalence. For each control, it gives Cohen's $d$ with a 90% confidence interval and applies the two one-sided tests (TOST) procedure against the schema's bound of half a standard deviation (Lakens, 2017). The frequency row reflects the intended manipulation, so Figure 1 leaves it out.

```{r}
report <- match_report(
  stimuli,
  c('frequency', 'length', 'n_density', 'old20'),
  schema
)
report$comparisons
```

```{r balance-plot, fig.width = 7.5, fig.height = 2.8, fig.cap = 'Realised Balance Between the Frequency Conditions on Each Control Variable, With 90% Confidence Intervals Against the Prespecified Equivalence Region. Shading marks the equivalence region set in the schema, from minus to plus half a standard deviation.', fig.alt = 'Dot-and-whisker plot of standardised differences, high minus low frequency, with 90 per cent confidence intervals for word length, neighbourhood density and OLD20. All three estimates lie near zero and all three intervals fall inside the shaded equivalence region from minus 0.5 to plus 0.5'}
controls <- subset(report$comparisons, dimension != 'frequency')
controls$dimension <- factor(
  controls$dimension,
  levels = rev(c('length', 'n_density', 'old20')),
  labels = rev(c('Word length', 'Neighbourhood density', 'OLD20'))
)
bound <- schema$equivalence$bound_d

ggplot(controls, aes(cohens_d, dimension)) +
  annotate('rect', xmin = -bound, xmax = bound, ymin = -Inf, ymax = Inf,
           fill = '#E8F1F5') +
  geom_vline(xintercept = 0, colour = '#66747B', linewidth = 0.5) +
  geom_pointrange(aes(xmin = d_ci_low, xmax = d_ci_high),
                  colour = '#006D77', linewidth = 0.8, size = 0.7) +
  scale_x_continuous(
    breaks = c(-bound, 0, bound), expand = expansion(0),
    labels = scales::label_number(accuracy = 0.1, style_negative = 'minus')
  ) +
  coord_cartesian(xlim = c(-1, 1) * bound * 1.12) +
  labs(x = 'Standardised difference (high minus low), 90% CI', y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.title.x = element_text(size = 10.5, margin = margin(t = 6)),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 11, hjust = 0, colour = 'grey20',
                                   margin = margin(r = 10)),
        plot.margin = margin(8, 10, 8, 8))
```

The intervals deserve more attention than the point estimates, because they carry the uncertainty. With few items per condition, an estimate near zero can still be too imprecise to support a claim of equivalence.

### Carry the Items Into the Experiment

The design also describes each trial, which presents a fixation cross, the word with a condition trigger locked to its onset, a response window and a blank interval of jittered length. The jitter is derived from a keyed hash of the design and seed, so rebuilding the experiment reproduces the same timing.

```{r}
design$events <- list(
  list(type = 'fixation', content = '+', duration_ms = 500),
  list(type = 'text', content = '{word}', duration_ms = 800,
       trigger = 'condition', onset_locked = TRUE),
  list(type = 'response', keys = c('f', 'j'), timeout_ms = 2000),
  list(type = 'blank', duration = list(jitter = c(400, 800), as = 'iti_ms'))
)

list_plan <- balance_lists(stimuli, design, schema)
trials <- counterbalance(stimuli, design, schema, list_plan$list_of_set)
trials <- resolve_trial_timing(trials, design, schema)
trials <- assign_triggers(trials)

c(cost_before = list_plan$report$cost_before,
  cost_after = list_plan$report$cost_after,
  swaps = list_plan$report$n_swaps)
head(trials[, c('list', 'trial', 'condition', 'word', 'iti_ms',
                'condition_trigger')])
```

The imbalance cost sums each list's departure from its fair share on every balanced variable, scaled to that variable's mean. Each swap exchanged a matched word pair in one list for a pair in the other.

All three exported files read from this trial table. In the PsychoPy script, `win.callOnFlip` sends the trigger on the refresh that presents the word, and `TRIGGER_HOLD_MS` sets how long the trigger code is held before the port is reset to zero.

```{r}
experiments <- export_experiments(trials, design, schema, output_dir)
basename(unlist(experiments))

psychopy_code <- readLines(experiments$psychopy, warn = FALSE)
grep('^\\s*(TRIGGER_HOLD_MS\\s*=|win\\.callOnFlip\\()', psychopy_code,
     value = TRUE, perl = TRUE)
```

## Discussion

As a diagnostic, the balance plot cannot show that the materials are theoretically adequate. A successful match can still leave a weak manipulation, a confounded norm or an implausible trial sequence. The workflow's value lies in making those choices visible before data collection, so that they can be reviewed alongside the exported files. In an EEG study, that review should extend to the trigger stream and hardware timing measured on the recording computer.

### Limits

Matching can only draw on what the candidate corpus contains. The bundled English lexicon takes its words and Zipf frequencies from wordfreq (Speer, 2022). It serves this demonstration, but it includes proper nouns, estimates syllables from spelling and has no part-of-speech information. A study should therefore supply its own exclusions and whatever norms it needs. Equivalence also depends on the bound chosen before selection, and the default of half a standard deviation is only a starting point. Each study should derive its own bound from the smallest difference that would matter for its research question (Lakens, 2017).

The [R documentation](https://pablobernabeu.github.io/lexsync/r/) and [Python documentation](https://pablobernabeu.github.io/lexsync/python/) cover pseudoword generation, semantic norms, continuous designs, Latin-square counterbalancing and the full file-based pipeline. Across the package, the design records the experiment as intended, while the generated tables and reports record it as realised.

```{r cleanup, include = FALSE}
unlink(output_dir, recursive = TRUE)
```

## References

Bernabeu, P. (2026). *lexsync: Lexical optimisation and hardware-timed experiment generation* (Version 0.1.0) [Computer software]. CRAN. https://doi.org/10.32614/CRAN.package.lexsync

Lakens, D. (2017). Equivalence tests: A practical primer for *t* tests, correlations, and meta-analyses. *Social Psychological and Personality Science*, *8*(4), 355–362. https://doi.org/10.1177/1948550617697177

Speer, R. (2022). *rspeer/wordfreq: v3.0* (Version 3.0.2) [Computer software]. Zenodo. https://doi.org/10.5281/zenodo.7199437

van Heuven, W. J. B., Mandera, P., Keuleers, E., & Brysbaert, M. (2014). SUBTLEX-UK: A new and improved word frequency database for British English. *Quarterly Journal of Experimental Psychology*, *67*(6), 1176–1190. https://doi.org/10.1080/17470218.2013.850521

Yarkoni, T., Balota, D., & Yap, M. (2008). Moving beyond Coltheart's N: A new measure of orthographic similarity. *Psychonomic Bulletin & Review*, *15*(5), 971–979. https://doi.org/10.3758/PBR.15.5.971
