---
title: 'lexsync: From a word-frequency corpus to an EEG-ready experiment'
author: ''
date: '2026-09-01'
slug: lexsync-from-corpus-to-eeg-ready-experiment
categories:
  - psycholinguistics
  - research methods
tags:
  - s
  - psycholinguistics
  - experimental stimuli
  - electroencephalography
  - reproducibility
  - Python
  - PsychoPy
  - OpenSesame
  - jsPsych
subtitle: ''
summary: 'Selecting words is only the first step in preparing a psycholinguistic experiment. This post uses lexsync to match high- and low-frequency items, balance them across lists and export presentation scripts with recorded timing and EEG triggers.'
abstract: 'This tutorial shows how lexsync turns a lexical design into checked stimulus lists and experiment files. It separates candidate selection from matching, evaluates balance on intended controls and carries the resulting item table into PsychoPy, OpenSesame or jsPsych with recorded timing and EEG trigger information.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 92)
xaringanExtra::use_clipboard()
output_dir <- file.path(tempdir(), 'lexsync-blog')
unlink(output_dir, recursive = TRUE)
dir.create(output_dir, showWarnings = FALSE)
```

## Introduction

A high- versus low-frequency word experiment sounds simple until the materials are assembled. The frequency bands must differ, while length and orthographic similarity should not. The selected words then need to be divided across presentation lists without creating a new imbalance. Finally, the same item table must reach the experiment software with its trial order, timing and electroencephalography (EEG) triggers intact.

I wrote [lexsync](https://github.com/pablobernabeu/lexsync) to keep those stages in one reproducible workflow. The package stores the design separately from the candidate corpus, reports the balance that was achieved and exports runnable experiments for PsychoPy, OpenSesame and jsPsych. The R and Python implementations use the same design format. This article uses R for brevity.

## Development

The workflow below keeps the design decisions visible: define the candidate pool, check the realised balance and export the same specification to the presentation environment.

```{r}
library(lexsync)
library(ggplot2)
packageVersion('lexsync')

schema_path <- system.file('extdata', 'schema.yaml', package = 'lexsync')
lexicon_path <- system.file('extdata', 'en_example.csv', package = 'lexsync')
schema <- yaml::read_yaml(schema_path)
lexicon <- load_lexicon(lexicon_path, schema, language = 'english')
```

### Put the design in data

The design below asks for 40 words in each frequency condition. Length, neighbourhood density and OLD20 are controls. The two lists will also be balanced on those controls and on frequency. Zipf frequency follows van Heuven et al. (2014), while OLD20 measures orthographic similarity as the mean edit distance from a word's 20 nearest neighbours (Yarkoni et al., 2008).

```{r}
design <- list(
  name = 'frequency_eeg',
  language = 'english',
  n_per_condition = 40,
  pool_filters = list(length = c(3, 7), frequency = c(3.8, 7.0)),
  conditions = list(
    list(name = 'high_frequency', define_by = list(frequency = c(5.2, 7.0))),
    list(name = 'low_frequency', define_by = list(frequency = c(3.8, 4.4)))
  ),
  match_on = c('length', 'n_density', 'old20'),
  counterbalance = list(
    lists = 2,
    optimise = TRUE,
    balance_on = c('length', 'n_density', 'old20', 'frequency')
  )
)

pool <- build_pool(lexicon, design$pool_filters)
candidate_counts <- vapply(
  design$conditions,
  function(condition) nrow(build_pool(pool, condition$define_by)),
  integer(1)
)
names(candidate_counts) <- vapply(design$conditions, `[[`, character(1), 'name')
candidate_counts
```

The high-frequency band is the smaller candidate set, so it limits the number and quality of possible matches. Keeping that count beside the design makes a shortfall interpretable instead of leaving it to appear as a mysterious matching failure.

### Select and check the words

`match_stimuli()` spreads anchor words across the smaller frequency band and finds nearby words from the other condition in the standardised space of the control variables.

```{r}
stimuli <- match_stimuli(pool, design, schema)
head(stimuli[, c('set', 'condition', 'word', 'frequency', 'length',
                  'n_density', 'old20')])
```

The report treats matching as an equivalence claim. For each control, it gives Cohen's $d$ with a 90% confidence interval and tests that interval against the design's bound of half a standard deviation (Lakens, 2017). The frequency row describes the intended manipulation and is omitted from the figure.

```{r}
report <- match_report(
  stimuli,
  c('frequency', 'length', 'n_density', 'old20'),
  schema
)
report$comparisons
```

```{r balance-plot, fig.width = 7.5, fig.height = 4.2, fig.alt = 'Dot-and-whisker plot showing standardised differences and 90 per cent confidence intervals for word length, neighbourhood density and OLD20, with equivalence bounds at minus and plus 0.5'}
controls <- subset(report$comparisons, dimension != 'frequency')
controls$dimension <- factor(
  controls$dimension,
  levels = rev(c('length', 'n_density', 'old20')),
  labels = rev(c('Word length', 'Neighbourhood density', 'OLD20'))
)

ggplot(controls, aes(cohens_d, dimension)) +
  annotate('rect', xmin = -0.5, xmax = 0.5, ymin = -Inf, ymax = Inf,
           fill = '#E8F1F5') +
  geom_vline(xintercept = 0, colour = '#66747B', linewidth = 0.5) +
  geom_pointrange(aes(xmin = d_ci_low, xmax = d_ci_high),
                  colour = '#006D77', linewidth = 0.7) +
  scale_x_continuous(limits = c(-0.65, 0.65), breaks = c(-0.5, 0, 0.5)) +
  labs(x = 'Standardised difference (low minus high), 90% CI', y = NULL,
       title = 'Realised balance on the control variables',
       subtitle = 'Shading marks the prespecified equivalence region') +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title.position = 'plot',
        plot.title = element_text(size = 11, face = 'bold'),
        plot.subtitle = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9))
```

The point estimates are useful, but the intervals carry the uncertainty. With only a few items, near-zero estimates can still be too imprecise to support an equivalence claim.

### Carry the items into the experiment

The design also describes a trial. A fixation is followed by the word, an onset-locked condition trigger, a response window and a jittered blank interval. Jitter is derived from a keyed hash of the design and seed, so rebuilding the experiment preserves the realised timing.

```{r}
design$events <- list(
  list(type = 'fixation', content = '+', duration_ms = 500),
  list(type = 'text', content = '{word}', duration_ms = 800,
       trigger = 'condition', onset_locked = TRUE),
  list(type = 'response', keys = c('f', 'j'), timeout_ms = 2000),
  list(type = 'blank', duration = list(jitter = c(400, 800), as = 'iti_ms'))
)

list_plan <- balance_lists(stimuli, design, schema)
trials <- counterbalance(stimuli, design, schema, list_plan$list_of_set)
trials <- resolve_trial_timing(trials, design, schema)
trials <- assign_triggers(trials)

c(cost_before = list_plan$report$cost_before,
  cost_after = list_plan$report$cost_after,
  swaps = list_plan$report$n_swaps)
head(trials[, c('list', 'trial', 'condition', 'word', 'iti_ms',
                'condition_trigger')])
```

The exported files all read from this trial table. In the PsychoPy script, `win.callOnFlip` sends the trigger on the refresh that presents the word.

```{r}
experiments <- export_experiments(trials, design, schema, output_dir)
basename(unlist(experiments))

psychopy_code <- readLines(experiments$psychopy, warn = FALSE)
grep('^\\s*(TRIGGER_HOLD_MS\\s*=|win\\.callOnFlip\\()', psychopy_code,
     value = TRUE, perl = TRUE)
```

## Discussion

The balance plot is a diagnostic, not a guarantee that the materials are theoretically adequate. A successful match can still leave a weak manipulation, a confounded norm or an implausible trial sequence. Its value is that those choices are visible before data collection and can be reviewed alongside the exported files. In an EEG study, that audit should be extended to the actual trigger stream and hardware timing on the recording computer.

### Limits

Matching can only use what the candidate corpus contains. The bundled lexicon is suitable for this demonstration, but it includes proper nouns, estimates syllables from spelling and has no part-of-speech information. A study should therefore supply its own exclusions and any norms needed for the research question. Equivalence also depends on the bound chosen before selection. The default half-standard-deviation region is not a universal definition of an acceptable match.

The [R documentation](https://pablobernabeu.github.io/lexsync/r/) and [Python documentation](https://pablobernabeu.github.io/lexsync/python/) cover pseudoword generation, semantic norms, continuous designs, Latin-square counterbalancing and the full file-based pipeline. The useful boundary is simple: the design records the intended experiment, while the generated tables and reports record what was realised.

```{r cleanup, include = FALSE}
unlink(output_dir, recursive = TRUE)
```

## References

Lakens, D. (2017). Equivalence tests: A practical primer for *t* tests, correlations, and meta-analyses. *Social Psychological and Personality Science*, *8*(4), 355–362. https://doi.org/10.1177/1948550617697177

van Heuven, W. J. B., Mandera, P., Keuleers, E., & Brysbaert, M. (2014). SUBTLEX-UK: A new and improved word frequency database for British English. *Quarterly Journal of Experimental Psychology*, *67*(6), 1176–1190. https://doi.org/10.1080/17470218.2013.850521

Yarkoni, T., Balota, D., & Yap, M. (2008). Moving beyond Coltheart's N: A new measure of orthographic similarity. *Psychonomic Bulletin & Review*, *15*(5), 971–979. https://doi.org/10.3758/PBR.15.5.971
