---
title: 'scopusflow: A literature search you can rerun'
author: ''
date: '2026-09-01'
slug: scopusflow-a-literature-search-you-can-rerun
categories:
  - bibliometrics
  - research methods
tags:
  - s
  - bibliometrics
  - literature review
  - reproducibility
  - open science
  - Python
  - Scopus
subtitle: ''
summary: 'A literature-search export does not preserve the decisions that produced it. This post uses scopusflow to define a search in advance, resume it from checkpoints and generate a PRISMA-S search record from the resulting data.'
abstract: 'This tutorial treats a Scopus search as a versioned, resumable workflow. It defines a query and a yearly plan, retrieves an offline demonstration corpus through checkpoints, compares DOI sets and generates a PRISMA-S search record that names the items left for the review authors to report.'
authors: []
lastmod: ''
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE,
                      fig.align = 'center', fig.retina = 2, comment = '#>')
options(width = 90, cli.num_colors = 1, cli.hyperlink = FALSE,
        crayon.enabled = FALSE)
xaringanExtra::use_clipboard()
```

## Introduction

A spreadsheet exported from Scopus contains papers, but it does not contain the search that found them. The exact query, retrieval date, paging choices and number of records returned by each part of the search often end up in notes or browser history. When the search has to be updated, it can be difficult to tell whether a changed result reflects the literature or the method.

I wrote [scopusflow](https://pablobernabeu.github.io/scopusflow/) (Bernabeu, 2026) to keep those decisions with the records. A search begins as a plan, each completed part is cached and the final object can write its own search record. Matching packages are available for R ([CRAN](https://CRAN.R-project.org/package=scopusflow), [GitHub](https://github.com/pablobernabeu/scopusflow)) and Python ([PyPI](https://pypi.org/project/scopusflow), [GitHub](https://github.com/pablobernabeu/scopusflow-py)).

The example below uses a small corpus bundled with the R package. A stand-in server, defined in [a short script](https://github.com/pablobernabeu/website-files/blob/master/content/post/scopusflow-a-literature-search-you-can-rerun/stand_in.R), answers the requests that would otherwise go to the Scopus API, so the code runs without an API key or any connection to Elsevier.

## Development

The workflow below separates planning, retrieval, inspection and reporting.

```{r}
library(scopusflow)
packageVersion('scopusflow')
```

### Describe the search first

The bundled example concerns research on graphene supercapacitors. `scopus_query()` adds the field syntax, and `scopus_plan()` divides the search into yearly cells. The plan can be inspected and saved before any quota is spent.

```{r}
query <- scopus_query('graphene supercapacitor', .field = 'TITLE-ABS-KEY')
plan <- scopus_plan(query, years = 2015:2024, partition = 'year')
query
plan
```

The plan already yields a draft methods description, written to PRISMA-S, the extension of the PRISMA statement for reporting literature searches (Rethlefsen et al., 2021). The draft also flags the items that the review authors will have to report themselves.

```{r}
cat(format(scopus_search_report(plan), style = 'paragraph'))
```

### Resume from the last completed cell

The bundled corpus now stands in for the result of that plan. The cache starts empty, so every yearly cell has to be requested.

```{r}
source('stand_in.R')
stand_in <- scopus_stand_in(example_records)
options(scopusflow.api_key = 'offline-demo-key')

cache <- file.path(tempdir(), 'scopusflow-blog-cache')
unlink(cache, recursive = TRUE)

httr2::with_mocked_responses(stand_in$handler, {
  records <- scopus_fetch_plan(plan, cache_dir = cache, resume = TRUE)
})
c(records = nrow(records), requests = stand_in$state$requests)
```

Each year is written to its own checkpoint. Repeating the plan reloads those files and makes no further request.

```{r}
requests_before <- stand_in$state$requests
httr2::with_mocked_responses(stand_in$handler, {
  repeated <- scopus_fetch_plan(plan, cache_dir = cache, resume = TRUE)
})
c(records = nrow(repeated),
  new_requests = stand_in$state$requests - requests_before)
```

The practical benefit is clearest after an interruption, since the completed cells are kept. Each checkpoint also stores the query, year, view and page size it was fetched under, which prevents an incompatible plan from silently reusing it.

### Inspect the result

The records carry their plan, retrieval time and package version, along with the count for each cell. That provenance later helps to tell a change in the literature apart from a change in the method. The plot below counts the bundled records for each year.

```{r records-per-year, fig.width = 7.5, fig.height = 4.2, fig.alt = 'Line chart showing the number of graphene-supercapacitor records in the bundled corpus for each year from 2015 to 2024'}
httr2::with_mocked_responses(stand_in$handler, {
  trend <- scopus_trend('graphene supercapacitor', years = 2015:2024,
                        field = 'TITLE-ABS-KEY')
})
plot_scopus_trend(trend) +
  ggplot2::labs(
    title = 'Records in the demonstration corpus',
    caption = 'Offline OpenAlex-derived demonstration records, 2015-2024'
  ) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11, face = 'bold'),
    axis.title = ggplot2::element_text(size = 10),
    axis.text = ggplot2::element_text(size = 9),
    plot.caption = ggplot2::element_text(size = 8.5, colour = 'grey35')
  )
```

Successive harvests can be compared by DOI. In this illustration, the baseline ends in 2023 and the later harvest adds the 2024 records.

```{r}
baseline <- records[records$year <= 2023, ]
changes <- scopus_diff_dois(old = baseline, new = records)
changes
```

### Let the result write the record

The search record is generated from the object that holds the publications. It reports what the software knows and identifies the PRISMA-S items that still require the review authors' judgement, such as citation searching or peer review of the search strategy. As this run never reached Scopus, the summary below leaves out the database, platform and retrieval date that a live report would include.

```{r report-summary, echo = FALSE}
report <- scopus_search_report(records)
knitr::kable(data.frame(
  Detail = c('Search expression', 'Years', 'Records in bundled corpus',
             'Records with a DOI', 'PRISMA-S checklist items'),
  Value = c(report$expression, report$years, report$n_records,
            report$n_with_doi, nrow(report$prisma))
), row.names = FALSE)
```

The same records can then leave R as a DOI list, BibTeX, RIS or a table for bibliometrix. Saving the native object preserves the attached provenance for a future update.

## Discussion

The result that matters in this example is the provenance. A rerunnable search records enough of the method to explain why a later DOI set differs, although database coverage, query design and screening remain matters of judgement. Those decisions belong in the review protocol and should be reported with the generated search record. For a review, I would keep the plan under version control, save the native result object and archive the search record alongside the screening data.

The [R reference](https://pablobernabeu.github.io/scopusflow/reference/) and the [Python documentation](https://pablobernabeu.github.io/scopusflow-py/) cover cursor paging, de-duplication, topic comparisons and authentication for live searches.

### Limits

The offline corpus only demonstrates the workflow, so its counts are not evidence about the contents of Scopus or about graphene research. Elsevier's terms do not allow Scopus records to be redistributed, which is why the example uses 138 records retrieved from OpenAlex on 22 July 2026 and reshaped to the package's return format. A live search still depends on Elsevier's coverage, indexing and quota, and abstract retrieval draws on a separate allowance.

```{r cleanup, include = FALSE}
options(scopusflow.api_key = NULL)
unlink(cache, recursive = TRUE)
```

## References

Bernabeu, P. (2026). *scopusflow: A reproducible workflow layer for Scopus bibliographic searches* (Version 0.4.0) [Computer software]. CRAN. https://doi.org/10.32614/CRAN.package.scopusflow

Rethlefsen, M. L., Kirtley, S., Waffenschmidt, S., Ayala, A. P., Moher, D., Page, M. J., Koffel, J. B., & PRISMA-S Group. (2021). PRISMA-S: An extension to the PRISMA statement for reporting literature searches in systematic reviews. *Systematic Reviews*, *10*(1), Article 39. https://doi.org/10.1186/s13643-020-01542-z
